### 2017-11-30 doc/doc.tex
documentation

\documentclass[a4paper,german]{report}

    \title{Projekt Medieninformatik}
    \date{WS17/18}
    \author{Lewe Ohlsen\\ Fachhochschule Wedel}

    \usepackage[utf8]{inputenc}                   % latin9 utf8
    \usepackage[german]{babel}
    \usepackage{tikz}
    \usepackage{pgfgantt}

    \begin{document}
        \maketitle
            \section{Section Headings}
            \begin{ganttchart}{1}{16}
                \gantttitle{Kalenderwoche}{16} \\
                \gantttitlelist{46,...,52,1,2,3,4,5,6,7,8,9}{1} \\

                \ganttbar{MNIST Tutorial}{1}{2} \\
                \ganttlinkedbar{Datenaufbereitung}{3}{3} \\
                \ganttbar{CNN-Inplementierung}{3}{9} \\

                %\ganttlinkedbar{Task 2}{3}{7} \ganttnewline
                \ganttmilestone{Webcam-Input}{7} \ganttnewline
                \ganttbar{Final Task}{8}{12}

                %\ganttlink{elem1}{elem2}

            \end{ganttchart}
    \end{document}

### 2018-01-10 doc/doc.tex
notes
chapters

@@
-\documentclass[a4paper,german]{report}
+\documentclass[a4paper,english]{report}
 
-    \title{Projekt Medieninformatik}
+    \title{Projekt Medieninformatik\\Facial Expresson Recognition\\ Using A State Of The Art CNN}
     \date{WS17/18}
@@
 
-    \usepackage[utf8]{inputenc}                   % latin9 utf8
-    \usepackage[german]{babel}
+    \usepackage[utf8]{inputenc}  
+    \usepackage[inline]{enumitem}
     \usepackage{tikz}
@@
         \maketitle
-            \section{Section Headings}
+        \newpage
+            \tableofcontents
+        \newpage
+        \section{Introduction}
+            In order to get started with deep learning techniques and state of the art neural networks, I am participating in a Facial Expression Recognition Challenge\footnote{https://www.kaggle.com/c/challenges-in-representation-learning-facial-expression-recognition-challenge/data} on Kaggle.
+The goal is to build a classifier for facial expression recognition on live images.
+        \section{Roadmap}
             \begin{ganttchart}{1}{16}
-                \gantttitle{Kalenderwoche}{16} \\
+                \gantttitle{Calendar Week}{16} \\
                 \gantttitlelist{46,...,52,1,2,3,4,5,6,7,8,9}{1} \\
@@
                 \ganttbar{MNIST Tutorial}{1}{2} \\
-                \ganttlinkedbar{Datenaufbereitung}{3}{3} \\
-                \ganttbar{CNN-Inplementierung}{3}{9} \\
+                \ganttlinkedbar{Data preparation}{3}{3} \\
+                \ganttbar{implementing a simple linear model}{3}{4} \\                                
+                \ganttlinkedbar{implementing a CNN}{4}{10} \\
 
                 %\ganttlinkedbar{Task 2}{3}{7} \ganttnewline
-                \ganttmilestone{Webcam-Input}{7} \ganttnewline
-                \ganttbar{Final Task}{8}{12}
+                \ganttmilestone{processing Webcam images}{8} \ganttnewline
+                \ganttlinkedbar{Tuning parameters}{10}{12}
 
@@
             \end{ganttchart}
+        \chapter{About the Dataset}
+        The Dataset consists of 48x48 pixel grayscale images of faces. The faces have been centered automatically do not require further preprocessing.
+        The Dataset is initially being split into two subsets: The model is first trained to fit the training data and subsequently evaluated using the test dataset.
+
+        % \begin{enumerate*}[label=(\roman*)]
+        %     \item Training,\label{item:training}
+        %     \item PublicTest and\label{item:pubtest}
+        %     \item PrivateTest.\label{item:prvtest}
+        % \end{enumerate*}\\
+        % \ref{item:training} Dogs are very nice animals. However they need to be walked several times a day\ldots [continuous long passage spreading over multiple lines]\\
+        % \ref{item:pubtest} Cats are often considered easier animals but\\
+        % \ref{item:prvtest} Cats are often considered easier animals but\ldots
+
+        \chapter{Feedforward Neural Networks}
+        \section{Perceptrons}
+        \section{Cost Function}
+        \section{Optimizers}
+        \section{Gradient Descend}
+        \section{Backpropagation}        
+        \chapter{Convolutional Neural Networks}
+        \section{Convolutions}
+        \section{Filters}
+        \section{Layers and Dropouts}
+        \chapter{Processing Live Images}
     \end{document}

### 2018-01-10 webcam.py
webcam images

@@ -1,0 +1,83 @@
+import face_recognition
+import cv2
+
+# This is a demo of running face recognition on live video from your webcam. It's a little more complicated than the
+# other example, but it includes some basic performance tweaks to make things run a lot faster:
+#   1. Process each video frame at 1/4 resolution (though still display it at full resolution)
+#   2. Only detect faces in every other frame of video.
+
+# PLEASE NOTE: This example requires OpenCV (the `cv2` library) to be installed only to read from your webcam.
+# OpenCV is *not* required to use the face_recognition library. It's only required if you want to run this
+# specific demo. If you have trouble installing it, try any of the other demos that don't require it instead.
+
+# Get a reference to webcam #0 (the default one)
+video_capture = cv2.VideoCapture(0)
+
+# Initialize some variables
+face_locations = []
+face_encodings = []
+face_emotions = []
+frame_count = 0
+process_every_n_frames = 10
+
+while True:
+    # Grab a single frame of video
+    ret, frame = video_capture.read()
+
+    # Resize frame of video to 1/4 size for faster face recognition processing
+    small_frame = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
+
+    # Convert the image from BGR color (which OpenCV uses) to RGB color (which face_recognition uses)
+    rgb_small_frame = small_frame[:, :, ::-1]
+
+    # Only process every other frame of video to save time
+    if frame_count % process_every_n_frames == 0:
+        # Find all the faces and face encodings in the current frame of video
+        face_locations = face_recognition.face_locations(rgb_small_frame)
+
+        face_emotions = []
+        for (top, right, bottom, left) in face_locations:
+            # Sacle face locations back up
+            top *= 4
+            right *= 4
+            bottom *= 4
+            left *= 4
+            # See if we can find out the facial expression
+            emotion = "Unknown"
+            # Write image to disk
+            face_image = frame[top:bottom, left:right]
+            cv2.imwrite("face%04i.jpg" %frame_count, face_image)
+
+            face_emotions.append(emotion)
+
+
+    # Display the results
+    for (top, right, bottom, left), name in zip(face_locations, face_emotions):
+        # Scale back up face locations since the frame we detected in was scaled to 1/4 size
+        top *= 4
+        right *= 4
+        bottom *= 4
[…]

### 2018-01-17 doc/doc.tex
challenges
4 layers

--- 2017-11-30 doc/doc.tex
@@
         \maketitle
-            \section{Section Headings}
+        \newpage
+            \tableofcontents
+        \newpage
+        \section{Introduction}
+            The projects goal is to build a device to process live webcam
+            images, detect faces and classify the emotion of each face.
+            In order to get started with deep learning techniques and state 
+            of the art neural networks, I am participating in a Facial 
+            Expression Recognition 
+            Challenge\footnote{https://www.kaggle.com/c/challenges-in-representation-learning-facial-expression-recognition-challenge/data} 
+            on Kaggle. 
+        \section{Roadmap}
             \begin{ganttchart}{1}{16}
-                \gantttitle{Kalenderwoche}{16} \\
+                \gantttitle{Calendar Week}{16} \\
                 \gantttitlelist{46,...,52,1,2,3,4,5,6,7,8,9}{1} \\
@@
             \end{ganttchart}
+        \chapter{About the Dataset}
+
+        The Dataset consists of 48x48 pixel grayscale images of faces. 
+        These images have been centered automatically to show a similar 
+        region of the face in every sample. All the images come classified
+        with the facial expression visible.
+        % TODO: explain what a model is
+        Initially the dataset is split into two subsets: A training set 
+        that is used to fit the model and a test set to evaluate the model
+        and see how well it performs.
+
+        Figure of classes and distribution…
+
+        \chapter{Challenges}
+        \section{Recognizing Faces}
+        \section{Classifying Facial Expressions}
+        \chapter{Feedforward Neural Networks}
+        A feedforward neural network can be thought of as a function f(x) = y
+        where x is a fixed-size vector that represents the input data. In this
+        case, the image is represented as a … dimensional vector, each component
+        being the grayscale intensity of a pixel [0..1]. y is a 7-dimensional vector,
+        each component representing the computed probability of the image belonging 
+        to a certain class. The function f is actually a computational graph (DAG)…
+
+        \section{Perceptrons}
+
+        \section{Cost Function}
+
+        \section{Optimizers}
+        \section{Gradient Descend}
+        \section{Backpropagation}        
+        \chapter{Convolutional Neural Networks}
+        \section{Convolutions}
+        \section{Filters}
+        \section{Layers and Dropouts}
+        \chapter{Processing Live Images}
     \end{document}

### 2018-01-20 predict.py
processing live images

@@ -1,0 +1,123 @@
+import tensorflow as tf
+import numpy as np
+import csv
+
+
+session = tf.Session()
+
+img_size = 48
+img_size_flat = img_size * img_size
+img_shape = (img_size, img_size)
+
+num_classes = 7
+class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
+dropout = 0.25
+
+imgs_test_byte = []
+
+with open('fer2013.csv') as csvfile:
+    readCSV = csv.reader(csvfile, delimiter=',')
+    # skip CSV header
+    next(readCSV)
+    for row in readCSV:
+        if row[2] == 'PrivateTest':
+            pixels_test = [float(x) for x in row[1].split(' ')]
+            imgs_test_byte.append(np.array(pixels_test))
+
+imgs_test_byte = np.array(imgs_test_byte, dtype=np.float32)
+imgs_test = np.divide(imgs_test_byte, 255)
+
+def conv_net(x_dict, n_classes, dropout, reuse, is_training):
+    
+    # Define a scope for reusing the variables
+    with tf.variable_scope('ConvNet', reuse=reuse):
+        # TF Estimator input is a dict, in case of multiple inputs
+        x = x_dict['images']
+
+        # The data input is a 1-D vector of 2304 features (48*48 pixels)
+        # Reshape to match picture format [Height x Width x Channel]
+        # Tensor input become 4-D: [Batch Size, Height, Width, Channel]
+        x = tf.reshape(x, shape=[-1, img_size, img_size, 1])
+
+        # Convolution Layer with 32 filters and a kernel size of 5
+        conv1 = tf.layers.conv2d(x, 32, 5, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv1 = tf.layers.max_pooling2d(conv1, 2, 2)
+
+        # Convolution Layer with 64 filters and a kernel size of 3
+        conv2 = tf.layers.conv2d(conv1, 64, 3, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv2 = tf.layers.max_pooling2d(conv2, 2, 2)
+        
+        # Convolution Layer with 128 filters and a kernel size of 3
+        conv3 = tf.layers.conv2d(conv2, 128, 3, activation=tf.nn.relu)
[…]

### 2018-01-20 webcam.py

@@ -1,5 +1,7 @@
 import face_recognition
 import cv2
+import numpy as np
+import predict
 
 # This is a demo of running face recognition on live video from your webcam. It's a little more complicated than the
 # other example, but it includes some basic performance tweaks to make things run a lot faster:
@@ -14,6 +16,7 @@ import cv2
 video_capture = cv2.VideoCapture(0)
 
 # Initialize some variables
+class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 face_locations = []
 face_encodings = []
 face_emotions = []
@@ -43,12 +46,18 @@ while True:
             bottom *= 4
             left *= 4
             # See if we can find out the facial expression
-            emotion = "Unknown"
-            # Write image to disk
-            face_image = frame[top:bottom, left:right]
-            cv2.imwrite("face%04i.jpg" %frame_count, face_image)
 
-            face_emotions.append(emotion)
+            face_image_rgb = frame[top:bottom, left:right]
+            face_image_gs = cv2.cvtColor(face_image_rgb, cv2.COLOR_BGR2GRAY)
+            face_image_gs_resized = cv2.resize(face_image_gs, (48, 48)) 
+            
+
+            img_byte = np.array(face_image_gs_resized, dtype=np.float32)
+            img = np.divide(img_byte, 255)
+
+            pred_class = predict.predictEmotion(img)
+
+            face_emotions.append(class_labels[pred_class])
 
 
     # Display the results

### 2018-01-22 doc/doc.tex
webcam processing, pretrained model

@@
         These images have been centered automatically to show a similar 
-        region of the face in every sample. All the images come classified
-        with the facial expression visible.
-        % TODO: explain what a model is
-        Initially the dataset is split into two subsets: A training set 
-        that is used to fit the model and a test set to evaluate the model
-        and see how well it performs.
+        region of the face in every sample. All the images are classified
+        with the vibible facial expression.
+\\\\
+Initially, the dataset is split into two subsets: A training set 
+        that is used to fit the prediction and a test set to evaluate the prediction results
+and compute the generalization error. The generalisation error is the average error for data that the model has never seen (e.g live webcam images).
 
-        Figure of classes and distribution…
+        \begin{figure}[h]
+            \centering
+            \includegraphics[width=0.6\textwidth]{figures/class_dist.png}
+            \caption{Class distribution across the training dataset}
+            \label{fig:classdist}
+        \end{figure}
 
@@
         \chapter{Feedforward Neural Networks}
-        A feedforward neural network can be thought of as a function f(x) = y
-        where x is a fixed-size vector that represents the input data. In this
-        case, the image is represented as a … dimensional vector, each component
+        A feedforward neural network can be thought of as a prediction function f(x) = y where x is a fixed-size vector that represents the input data. In this
+        case, the image is represented as a 2304-dimensional vector, each component
         being the grayscale intensity of a pixel [0..1]. y is a 7-dimensional vector,
@@
         to a certain class. The function f is actually a computational graph (DAG)…
+
+        The flow of information is uni-directional and not cyclic.
 
@@
         \section{Convolutions}
+    	\section{Pooling}
+    	Downsampling the image or hidden-layer output matrix and providing an abstracted form of the representation.
+This allows for assumptions to be made about features contained in the sub-regions of the image.
+
+    	Max-pooling e.g a 2x2 filter running over the image with a stride of 2 and taking the max value for each position.
+
+   		\begin{figure}[h]
+    		\centering
+    		\includegraphics[width=0.6\textwidth]{figures/max_pooling.png}
+    		\caption{Max-pooling a 4x4-pixels image with a 2x2 filter and a stride of 2}
+    		\label{fig:maxpooling}
+    	\end{figure}
+
+    	(Source: Stanford's CS231n github)
+
         \section{Filters}

### 2018-01-22 predict.py

@@ -54,8 +54,13 @@ def conv_net(x_dict, n_classes, dropout, reuse, is_training):
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
         conv3 = tf.layers.max_pooling2d(conv3, 2, 2)
 
+        # Convolution Layer with 256 filters and a kernel size of 3
+        conv4 = tf.layers.conv2d(conv3, 256, 3, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv4 = tf.layers.max_pooling2d(conv4, 2, 2)
+
         # Flatten the data to a 1-D vector for the fully connected layer
-        fc1 = tf.contrib.layers.flatten(conv3)
+        fc1 = tf.contrib.layers.flatten(conv4)
 
         # Fully connected layer (in tf contrib folder for now)
         fc1 = tf.layers.dense(fc1, 1024)
@@ -117,7 +122,5 @@ def predictEmotion(img):
     # Use the model to predict the images class
     preds = list(model.predict(input_fn))
 
-    print(preds)
+    print(preds, " -> ", class_labels[preds[0]])
     return preds[0]
-
-predictEmotion(imgs_test)

### 2018-01-22 webcam.py

@@ -40,11 +40,12 @@ while True:
 
         face_emotions = []
         for (top, right, bottom, left) in face_locations:
-            # Sacle face locations back up
+            # Scale face locations back up
             top *= 4
             right *= 4
             bottom *= 4
             left *= 4
+            
             # See if we can find out the facial expression
 
             face_image_rgb = frame[top:bottom, left:right]

### 2018-02-21 predict.py
vanilla opencv face detection

@@ -13,20 +13,6 @@ num_classes = 7
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 dropout = 0.25
 
-imgs_test_byte = []
-
-with open('fer2013.csv') as csvfile:
-    readCSV = csv.reader(csvfile, delimiter=',')
-    # skip CSV header
-    next(readCSV)
-    for row in readCSV:
-        if row[2] == 'PrivateTest':
-            pixels_test = [float(x) for x in row[1].split(' ')]
-            imgs_test_byte.append(np.array(pixels_test))
-
-imgs_test_byte = np.array(imgs_test_byte, dtype=np.float32)
-imgs_test = np.divide(imgs_test_byte, 255)
-
 def conv_net(x_dict, n_classes, dropout, reuse, is_training):
     
     # Define a scope for reusing the variables

### 2018-02-21 webcam.py

@@ -1,59 +1,48 @@
-import face_recognition
 import cv2
 import numpy as np
 import predict
 
-# This is a demo of running face recognition on live video from your webcam. It's a little more complicated than the
-# other example, but it includes some basic performance tweaks to make things run a lot faster:
-#   1. Process each video frame at 1/4 resolution (though still display it at full resolution)
-#   2. Only detect faces in every other frame of video.
-
-# PLEASE NOTE: This example requires OpenCV (the `cv2` library) to be installed only to read from your webcam.
-# OpenCV is *not* required to use the face_recognition library. It's only required if you want to run this
-# specific demo. If you have trouble installing it, try any of the other demos that don't require it instead.
-
 # Get a reference to webcam #0 (the default one)
-video_capture = cv2.VideoCapture(0)
+webcam = cv2.VideoCapture(0)
 
-# Initialize some variables
+# load OpenCV pre-trained cascade classifier for face detection
+face_cascade = cv2.CascadeClassifier('model/haarcascade_frontalface_default.xml')
+
+# variables
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 face_locations = []
-face_encodings = []
 face_emotions = []
 frame_count = 0
 process_every_n_frames = 10
 
 while True:
     # Grab a single frame of video
-    ret, frame = video_capture.read()
+    ret, frame = webcam.read()
 
     # Resize frame of video to 1/4 size for faster face recognition processing
-    small_frame = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
-
-    # Convert the image from BGR color (which OpenCV uses) to RGB color (which face_recognition uses)
-    rgb_small_frame = small_frame[:, :, ::-1]
+    # frame_resized = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
 
     # Only process every other frame of video to save time
     if frame_count % process_every_n_frames == 0:
-        # Find all the faces and face encodings in the current frame of video
-        face_locations = face_recognition.face_locations(rgb_small_frame)
+
+        # convert image to grayscale
+        frame_gray = cv2.cvtColor(frame, cv2.COLOR_BGR2GRAY)
+
+        # Find all the faces in the current frame of video
[…]

### 2018-02-21 webcam.py
quit with q

@@ -73,5 +73,5 @@ while True:
         break
 
 # Release handle to the webcam
-video_capture.release()
+webcam.release()
 cv2.destroyAllWindows()

### 2018-02-21 predict.py
resizing images in favor of performance

@@ -1,7 +1,5 @@
 import tensorflow as tf
 import numpy as np
-import csv
-
 
 session = tf.Session()
 

### 2018-02-21 webcam.py

@@ -20,38 +20,43 @@ while True:
     ret, frame = webcam.read()
 
     # Resize frame of video to 1/4 size for faster face recognition processing
-    # frame_resized = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
+    frame_resized = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
 
     # Only process every other frame of video to save time
     if frame_count % process_every_n_frames == 0:
 
         # convert image to grayscale
-        frame_gray = cv2.cvtColor(frame, cv2.COLOR_BGR2GRAY)
+        frame_resized_gray = cv2.cvtColor(frame_resized, cv2.COLOR_BGR2GRAY)
 
         # Find all the faces in the current frame of video
-        faces = face_cascade.detectMultiScale(frame_gray, 1.1, 5)
+        faces = face_cascade.detectMultiScale(frame_resized_gray, 1.1, 5)
 
         print('Faces found: ', len(faces))
 
         face_emotions = []
         for (x, y, w, h) in faces:
+
             print( x, y, w, h)
 
-            face_rgb = frame[y:y+h, x:x+w]
-            face_gs = cv2.cvtColor(face_rgb, cv2.COLOR_BGR2GRAY)
-            face_gs_resized = cv2.resize(face_gs, (48, 48))
-            
+            face = frame_resized_gray[y:y+h, x:x+w]
+            face_input = cv2.resize(face, (48, 48))
+            face_input_byte = np.array(face_input, dtype=np.float32)
+            face_input_intensity = np.divide(face_input_byte, 255)
 
-            img_byte = np.array(face_gs_resized, dtype=np.float32)
-            img = np.divide(img_byte, 255)
+            print(face_input)
 
-            pred_class = predict.predictEmotion(img)
+            pred_class = predict.predictEmotion(face_input_intensity)
 
             face_emotions.append(class_labels[pred_class])
 
 
     # Display the results
     for (x, y, w, h), name in zip(faces, face_emotions):
+
+        x *= 4
+        y *= 4
+        w*= 4
+        h *= 4
 
         # Draw a box around the face
         cv2.rectangle(frame,(x,y),(x+w,y+h),(0,0,255),2)

### 2018-02-21 model.py
separate models: model, train and predict

@@ -1,0 +1,100 @@
+import tensorflow as tf
+
+
+# Image dimensions
+img_size = 48
+img_size_flat = img_size * img_size
+img_shape = (img_size, img_size)
+
+# Class info
+num_classes = 7
+class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
+
+# Hyperparameters
+learning_rate = 0.001
+num_steps = 20000
+batch_size = 128
+dropout = 0.25
+
+
+def conv_net(x_dict, n_classes, dropout, reuse, is_training):
+    
+    # Define a scope for reusing the variables
+    with tf.variable_scope('ConvNet', reuse=reuse):
+        # TF Estimator input is a dict, in case of multiple inputs
+        x = x_dict['images']
+
+        # The data input is a 1-D vector of 2304 features (48*48 pixels)
+        # Reshape to match picture format [Height x Width x Channel]
+        # Tensor input become 4-D: [Batch Size, Height, Width, Channel]
+        x = tf.reshape(x, shape=[-1, img_size, img_size, 1])
+
+        # Convolution Layer with 32 filters and a kernel size of 5
+        conv1 = tf.layers.conv2d(x, 32, 5, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv1 = tf.layers.max_pooling2d(conv1, 2, 2)
+
+        # Convolution Layer with 64 filters and a kernel size of 3
+        conv2 = tf.layers.conv2d(conv1, 64, 3, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv2 = tf.layers.max_pooling2d(conv2, 2, 2)
+        
+        # Convolution Layer with 128 filters and a kernel size of 3
+        conv3 = tf.layers.conv2d(conv2, 128, 3, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv3 = tf.layers.max_pooling2d(conv3, 2, 2)
+
+        # Convolution Layer with 256 filters and a kernel size of 3
+        conv4 = tf.layers.conv2d(conv3, 256, 3, activation=tf.nn.relu)
+        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
+        conv4 = tf.layers.max_pooling2d(conv4, 2, 2)
+
+        # Flatten the data to a 1-D vector for the fully connected layer
[…]

### 2018-02-21 predict.py

@@ -1,101 +1,11 @@
 import tensorflow as tf
 import numpy as np
+import model
 
 session = tf.Session()
 
-img_size = 48
-img_size_flat = img_size * img_size
-img_shape = (img_size, img_size)
-
-num_classes = 7
-class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
-dropout = 0.25
-
-def conv_net(x_dict, n_classes, dropout, reuse, is_training):
-    
-    # Define a scope for reusing the variables
-    with tf.variable_scope('ConvNet', reuse=reuse):
-        # TF Estimator input is a dict, in case of multiple inputs
-        x = x_dict['images']
-
-        # The data input is a 1-D vector of 2304 features (48*48 pixels)
-        # Reshape to match picture format [Height x Width x Channel]
-        # Tensor input become 4-D: [Batch Size, Height, Width, Channel]
-        x = tf.reshape(x, shape=[-1, img_size, img_size, 1])
-
-        # Convolution Layer with 32 filters and a kernel size of 5
-        conv1 = tf.layers.conv2d(x, 32, 5, activation=tf.nn.relu)
-        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
-        conv1 = tf.layers.max_pooling2d(conv1, 2, 2)
-
-        # Convolution Layer with 64 filters and a kernel size of 3
-        conv2 = tf.layers.conv2d(conv1, 64, 3, activation=tf.nn.relu)
-        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
-        conv2 = tf.layers.max_pooling2d(conv2, 2, 2)
-        
-        # Convolution Layer with 128 filters and a kernel size of 3
-        conv3 = tf.layers.conv2d(conv2, 128, 3, activation=tf.nn.relu)
-        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
-        conv3 = tf.layers.max_pooling2d(conv3, 2, 2)
-
-        # Convolution Layer with 256 filters and a kernel size of 3
-        conv4 = tf.layers.conv2d(conv3, 256, 3, activation=tf.nn.relu)
-        # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
-        conv4 = tf.layers.max_pooling2d(conv4, 2, 2)
-
-        # Flatten the data to a 1-D vector for the fully connected layer
[…]

### 2018-02-21 train.py

@@ -1,0 +1,62 @@
+import numpy as np
+import tensorflow as tf
+import csv
+
+import model
+
+# Feature Vectors (0-255)
+imgs_train_byte = []
+imgs_test_byte = []
+
+# Labels (0-6)
+labels_train_class = []
+labels_test_class = []
+
+
+with open('fer2013.csv') as csvfile:
+    readCSV = csv.reader(csvfile, delimiter=',')
+    # skip CSV header
+    next(readCSV)
+    for row in readCSV:
+        if row[2] == 'Training':
+            # cast pixels to int
+            pixels_train = [float(x) for x in row[1].split(' ')]
+            imgs_train_byte.append(np.array(pixels_train))
+            labels_train_class.append(int(row[0]))
+        elif row[2] == 'PrivateTest':
+            pixels_test = [float(x) for x in row[1].split(' ')]
+            imgs_test_byte.append(np.array(pixels_test))
+            labels_test_class.append(int(row[0]))
+
+
+# cast to correct type and wrap into np arrays
+imgs_train_byte = np.array(imgs_train_byte, dtype=np.float32)
+imgs_test_byte = np.array(imgs_test_byte, dtype=np.float32)
+labels_train_class = np.array(labels_train_class, dtype=np.float32)
+labels_test_class = np.array(labels_test_class, dtype=np.float32)
+# normalize the pixel intensitys
+imgs_train = np.divide(imgs_train_byte, 255)
+imgs_test = np.divide(imgs_test_byte, 255)
+
+
+# Build the Estimator
+estimator = tf.estimator.Estimator(model_fn=model.model_fn, model_dir='../model')
+
+
+# Training
+# Define the input function for training
+input_fn = tf.estimator.inputs.numpy_input_fn(
+    x={'images': imgs_train}, y=labels_train_class,
+    batch_size=model.batch_size, num_epochs=None, shuffle=True)
+# Train the Model
+estimator.train(input_fn, steps=model.num_steps)
+
+
+
+# Evaluate the Model
+# Define the input function for evaluating
+input_fn = tf.estimator.inputs.numpy_input_fn(
+    x={'images': imgs_test}, y=labels_test_class,
+    batch_size=model.batch_size, shuffle=False)
+# Use the Estimator 'evaluate' method
+estimator.evaluate(input_fn)

### 2018-02-21 webcam.py

@@ -16,7 +16,7 @@ frame_count = 0
 process_every_n_frames = 10
 
 while True:
-    # Grab a single frame of video
+    # Grab a single frame of the video
     ret, frame = webcam.read()
 
     # Resize frame of video to 1/4 size for faster face recognition processing
@@ -36,17 +36,14 @@ while True:
         face_emotions = []
         for (x, y, w, h) in faces:
 
-            print( x, y, w, h)
+            print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
             face_input = cv2.resize(face, (48, 48))
             face_input_byte = np.array(face_input, dtype=np.float32)
             face_input_intensity = np.divide(face_input_byte, 255)
 
-            print(face_input)
-
             pred_class = predict.predictEmotion(face_input_intensity)
-
             face_emotions.append(class_labels[pred_class])
 
 

### 2018-02-22 doc/doc.tex
ideas

@@
         \newpage
-        \section{Introduction}
+
+        \section{Abstract}
+            Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI).
             The projects goal is to build a device to process live webcam
@@
             on Kaggle.
+
+            As Convolutional Neural Networks (CNNs) have proven to be successful with many kinds of image classification  problems, I will focus on building a classifier using a state of the art CNN implementation.
         \section{Roadmap}
@@
         These images have been centered automatically to show a similar 
-        region of the face in every sample. All the images are classified
-        with the vibible facial expression.
-\\\\
+        region of the face in every sample. All the images are labeled
+        with the visible facial expression according to Ekmans (…) face classification (1970s).
+
         Initially, the dataset is split into two subsets: A training set 
-        that is used to fit the prediction and a test set to evaluate the prediction results
-and compute the generalization error. The generalisation error is the average error for data that the model has never seen (e.g live webcam images).
+        that is used to fit the prediction and a test set to evaluate the prediction results and compute the generalization error.
+The generalisation error is the average error for data that the model has never seen (e.g live webcam images).
 
@@
         \chapter{Convolutional Neural Networks}
+        %\section{Feature Extraction}
         \section{Convolutions}

### 2018-02-22 doc/doc.tex
new paper

@@
 
-        \section{Abstract}
-            Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI).
+        \section{Introduction}
+            Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI) and artificial emotional intelligence.
             The projects goal is to build a device to process live webcam

### 2018-02-22 doc/paper.tex
new paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Facial Expresion Recognition with Convolutional Neural Networks
% 
% Lewe Ohlsen
% Fachhochschule Wedel
%
% Latex Template by Frits Wenneker (http://www.howtotex.com)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[twoside]{article}

\usepackage{lipsum} % Package to generate dummy text throughout this template
\usepackage{graphicx} % Figures from matplotlib
\graphicspath{ {figures/} }

\usepackage[sc]{mathpazo} % Use the Palatino font
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures

\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text
\usepackage{paralist} % Used for the compactitem environment which makes bullet points with less space between them

\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

\usepackage{titlesec} % Allows customization of titles
\usepackage[utf8]{inputenc}
\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\title{\vspace{-15mm}\fontsize{16pt}{10pt}\selectfont\textbf{Facial Expression Recognition with\\ Convolutional Neural Networks}} % Article title

\author{
	\large
	\textsc{Lewe Ohlsen}\\[2mm] % Your name
	\normalsize Fachhochschule Wedel \\ % Your institution
	\normalsize \href{mailto:minf101062@fh-wedel.de}{minf101062@fh-wedel.de} % Your email address
	\vspace{-5mm}
}
\date{}

%----------------------------------------------------------------------------------------

\begin{document}

\maketitle % Insert title

%----------------------------------------------------------------------------------------
%	ABSTRACT
%----------------------------------------------------------------------------------------

\begin{abstract}

\noindent El barómetro es un instrumento de medición atmosérica, específicamente utilizado en la determinación de la fuerza por unidad de superficie ejercida por el peso de la atmósfera. Existe un gran número de equipos atmoféricos con distintos tipos de estos aparátos y son diariamente utilizados ya que la presión atmosférica juega un papel importante en la determinación y pronóstico del tiempo así como en el área de investigación al momento de realizarse experimentos ya que pueden llegar a afectar o hacer variar el funcionamiento de muchos aparatos electrónicos y mecánicos.

\end{abstract}

%----------------------------------------------------------------------------------------
%	ARTICLE CONTENTS
%----------------------------------------------------------------------------------------

\begin{multicols}{2} % Two-column layout throughout the main article text

\section{Introduction}

Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI) and artificial emotional intelligence. The projects goal is to build a device to process live webcam images, detect faces and classify the emotion of each face.

As Convolutional Neural Networks (CNNs) have proven to be successful with many kinds of image classification  problems, I will focus on building a classifier using a state of the art CNN implementation.

%------------------------------------------------

\section{Datasets}

Paul Ekmans 1969 research paper "Facial Expressions" elaborates that facial expressions can be interpreted as emotions. He suggests that there are six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise. In this project, these classes are used as emotion classes for predictions.

For training an artificial neural network to classify facial expressions to these basic emotions, a dataset with labeled images of faces is required.

\subsection{FER-2013}
The FER-2013 Dataset consists of roughly 37.000 images showing faces of people expressing a certain emotion. Images are 48x48 pixel grayscale images of faces. These images have been centered automatically to show a similar region of the face in every sample.

Initially, the dataset is split into two subsets: A training set that is used to fit the prediction and a test set to evaluate the prediction results.

\begin{figure}[H]
	\includegraphics[width=0.48\textwidth]{class_dist}
	\caption{Class distribution across the training dataset}
\end{figure}

\subsection{Cohn-Kanade}

%------------------------------------------------

\section{Discussion and Conclusion}

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

\begin{thebibliography}{99} % Bibliography - this is intentionally simple in this template

\bibitem[http://fluidos.eia.edu.co/]{Figueredo:2009dg}

\newblock {\em Fluidos, Carlos Toro s.}

\bibitem[http://www.windows2universe.org/]{Figueredo:2009dg}

\newblock {\em Becca Hatheway}

\bibitem[http://meteorologia.pucp.edu.pe/]{Figueredo:2009dg}

\newblock {\em Hernan Castillo, Pedro Ríos}

\bibitem[http://www.ammonit.com/Manuales]{Figueredo:2009dg}

\newblock {\em Ammonit Measurement GmbH}

\end{thebibliography}

%----------------------------------------------------------------------------------------

\end{multicols}

\end{document}

### 2018-02-28 model.py
one more layer, summaries

@@ -16,6 +16,8 @@ num_steps = 20000
 batch_size = 128
 dropout = 0.25
 
+# Enable logging
+tf.logging.set_verbosity(tf.logging.INFO)
 
 def conv_net(x_dict, n_classes, dropout, reuse, is_training):
     
@@ -30,24 +32,29 @@ def conv_net(x_dict, n_classes, dropout, reuse, is_training):
         x = tf.reshape(x, shape=[-1, img_size, img_size, 1])
 
         # Convolution Layer with 32 filters and a kernel size of 5
-        conv1 = tf.layers.conv2d(x, 32, 5, activation=tf.nn.relu)
+        conv1 = tf.layers.conv2d(x, 32, 5, padding='same', activation=tf.nn.relu)
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
         conv1 = tf.layers.max_pooling2d(conv1, 2, 2)
 
         # Convolution Layer with 64 filters and a kernel size of 3
-        conv2 = tf.layers.conv2d(conv1, 64, 3, activation=tf.nn.relu)
+        conv2 = tf.layers.conv2d(conv1, 64, 3, padding='same', activation=tf.nn.relu)
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
         conv2 = tf.layers.max_pooling2d(conv2, 2, 2)
         
         # Convolution Layer with 128 filters and a kernel size of 3
-        conv3 = tf.layers.conv2d(conv2, 128, 3, activation=tf.nn.relu)
+        conv3 = tf.layers.conv2d(conv2, 128, 3, padding='same', activation=tf.nn.relu)
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
         conv3 = tf.layers.max_pooling2d(conv3, 2, 2)
 
         # Convolution Layer with 256 filters and a kernel size of 3
-        conv4 = tf.layers.conv2d(conv3, 256, 3, activation=tf.nn.relu)
+        conv4 = tf.layers.conv2d(conv3, 256, 3, padding='same', activation=tf.nn.relu)
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
         conv4 = tf.layers.max_pooling2d(conv4, 2, 2)
+
+        # Convolution Layer with 256 filters and a kernel size of 3
+        conv5 = tf.layers.conv2d(conv4, 256, 3, padding='same', activation=tf.nn.relu)
[…]

### 2018-02-28 train.py
directories

@@ -13,7 +13,7 @@ labels_train_class = []
 labels_test_class = []
 
 
-with open('fer2013.csv') as csvfile:
+with open('data/fer2013.csv') as csvfile:
     readCSV = csv.reader(csvfile, delimiter=',')
     # skip CSV header
     next(readCSV)
@@ -40,7 +40,7 @@ imgs_test = np.divide(imgs_test_byte, 255)
 
 
 # Build the Estimator
-estimator = tf.estimator.Estimator(model_fn=model.model_fn, model_dir='../model')
+estimator = tf.estimator.Estimator(model_fn=model.model_fn, model_dir='./model')
 
 
 # Training

### 2018-03-03 doc/paper.tex
ckplus elaborations

@@
 
-\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text
 \usepackage{paralist} % Used for the compactitem environment which makes bullet points with less space between them
@@
 \usepackage[utf8]{inputenc}
-\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
-\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
-\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
-\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{} % Change the look of the section titles
+%\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
+%\renewcommand\thesubsection{\Roman{subsection}} % Roman numerals for subsections
+\titleformat{\section}[block]{\large\scshape\centering\bfseries}{\thesection.}{1em}{} % Change the look of the section titles
+\titleformat{\subsection}[block]{}{\thesubsection.}{1em}{} % Change the look of the section titles
 
@@
 
-Paul Ekmans 1969 research paper "Facial Expressions" elaborates that facial expressions can be interpreted as emotions.
-He suggests that there are six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise.
-In this project, these classes are used as emotion classes for predictions.
+Paul Ekmans 1969 research paper "Facial Expressions" elaborates that facial expressions can be interpreted and categorized into six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise.
+In this project, these classes are used for training and predictions.
 
-For training an artificial neural network to classify facial expressions to these basic emotions, a dataset with labeled images of faces is required.
+To build a classifier for facial expressions, a dataset with labeled images of faces is required.
+Each of the datasets is split into two subsets: A training set that is used to fit the prediction and a test set to evaluate the prediction results.
 
 \subsection{FER-2013}
+
 The FER-2013 Dataset consists of roughly 37.000 images showing faces of people expressing a certain emotion.
-Images are 48x48 pixel grayscale images of faces.
-These images have been centered automatically to show a similar region of the face in every sample.
-
-Initially, the dataset is split into two subsets: A training set that is used to fit the prediction and a test set to evaluate the prediction results.
+Most of the 48x48 pixel grayscale images are stock footage.
+These images have been centered on the persons faces to show a similar region in every sample.
+As this has been done automatically, some images are not centered correctly.
+Faces are snapped from very different angles, hence generalization of these samples is difficult.
 
@@
 	\includegraphics[width=0.48\textwidth]{class_dist}
-	\caption{Class distribution across the training dataset}
+	\caption{FER-2013 emotion class distribution}
 \end{figure}
 
-\subsection{Cohn-Kanade}
+\subsection{Cohn-Kanade (CK+)}
+To argument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
+CK+ consists of 593 image sequences across 123 subjects.
+Each image sequence goes from the neutral face to a "peak expression" according to the six basic emotions.
 

### 2018-03-03 util/ckplus_to_csv.py
processing ckplus images into a single csv file

@@ -1,0 +1,70 @@
+#!/usr/bin/env python
+
+"""This module reads the CK+ dataset emotion labels and the corresponding
+png images into a CSV file for easier processing in higher level apis.
+All images are converted to 48x48 pixels grayscale.
+
+To run this module, you should have the same folder structure as shown in 
+README.md.
+
+Labels are:
+    0=neutral
+    1=anger
+    2=contempt
+    3=disgust
+    4=fear
+    5=happy
+    6=sadness
+    7=surprise
+"""
+
+import os
+import glob
+import cv2
+
+# csv file to write to
+csv_file = open('../data/ckplus.csv', 'w')
+
+for root, dirs, files in os.walk("../data/ckplus/labels"):
+    for file in files:
+        csv_line = ''
+        label_path = os.path.join(root, file)
+        # read labels
+        with open(label_path, 'r') as f:
+            label = int(float(f.read()))
+            csv_line += str(label) + ','
+            # print(label)
+        # get images for label
+        image_paths = root.replace('labels', 'images')
+        image_paths = sorted(glob.glob(image_paths + '/*.png'))
+        # last image is the peak expression
+        peak_image = cv2.imread(image_paths[-1], flags=cv2.IMREAD_COLOR)
+        neutral_image = cv2.imread(image_paths[0], flags=cv2.IMREAD_COLOR)
+
+        # convert images to grayscale
+        peak_image = cv2.cvtColor(peak_image, cv2.COLOR_BGR2GRAY)
+        
+        if (label == 2):
+            cv2.imshow('Neutral Image', neutral_image)
+            cv2.imshow('Peak Image', peak_image)
+            cv2.waitKey()
+
+        # load OpenCV pre-trained cascade classifier for face detection
+        face_cascade = cv2.CascadeClassifier('../model/haarcascade_frontalface_default.xml')
+
+        # find and extract faces on image (should be one face per image)
+        # assuming all the falsely detected faces are smaller than the truly detected
+        # face, so discard all location params except of the largest
+        faces = face_cascade.detectMultiScale(peak_image, 1.1, 5)
+        (x, y, w, h) =  sorted(faces, key=len)[0]
+        # print(x, y, w, h)
+        face = peak_image[y:y+h, x:x+w]
+        # resize to 48x48 pixels
+        face = cv2.resize(face, (48, 48))
+
+        # write class and image pixels to csv
[…]

### 2018-03-03 doc/paper.tex
ckplus elaborations

@@
 
-Paul Ekmans 1969 research paper "Facial Expressions" elaborates that facial expressions can be interpreted and categorized into six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise.
+Paul Ekmans 1969 research paper "Facial Expressions" (REF) elaborates that facial expressions can be interpreted and categorized into six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise.
 In this project, these classes are used for training and predictions.
@@
 To argument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
-CK+ consists of 593 image sequences across 123 subjects.
-Each image sequence goes from the neutral face to a "peak expression" according to the six basic emotions.
+The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects.
+Each image sequence goes from a neutral facial expression to a "peak expression" according to the six basic emotions.
+Because only 327 of the 593 image sequences are labeled with emotion data.
+This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS) according to Ekman et al.
+CK+ also contains FACS data, in which action units (e.g raised eyebrow) are identified along with their degree of activation.
+
+Figure of CK+ distribution with neutral images.
 

### 2018-03-03 model.py
logging for tensorboard summary

@@ -21,6 +21,8 @@ def conv_net(x_dict, n_classes, dropout, reuse, is_training):
         # Tensor input become 4-D: [Batch Size, Height, Width, Channel]
         x = tf.reshape(x, shape=[-1, img_size, img_size, 1])
 
+        tf.summary.image("img", x)
+
         # Convolution Layer with 32 filters and a kernel size of 5
         conv1 = tf.layers.conv2d(x, 32, 5, padding='same', activation=tf.nn.relu)
         # Max Pooling (down-sampling) with strides of 2 and kernel size of 2
@@ -86,8 +88,8 @@ def model_fn(features, labels, mode, params):
     acc_op = tf.metrics.accuracy(labels=labels, predictions=pred_classes)
 
     # TODO: Write summary for tensorboard
-    # tf.summary.scalar("loss", loss_op)
-    # tf.summary.scalar("accuracy", acc_op)
+    tf.summary.scalar("loss", loss_op)
+    tf.summary.scalar("accuracy", acc_op[0])
     
     # TF Estimators requires to return a EstimatorSpec, that specify
     # the different ops for training, evaluating, ...

### 2018-03-03 predict.py
testing ckplus predictions - not too bad :)

@@ -7,10 +7,11 @@ class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral
 # Build the Estimator
 estimator = tf.estimator.Estimator(
     model_fn=model.model_fn, 
-    model_dir='./model',
+    model_dir='./model/ckplus',
     params={
         'learning_rate': 0.001,
         'num_classes': 7,
+        'img_size': 128,
         'dropout_rate': 0.25
     })
 

### 2018-03-03 doc/paper.tex
bibliography

@@
 
-\begin{thebibliography}{99} % Bibliography - this is intentionally simple in this template
+\begin{thebibliography}{9}
 
-\bibitem[http://fluidos.eia.edu.co/]{Figueredo:2009dg}
+	% OpenCV Haarcascade Classifier
+	\bibitem{violajones}
+  	P. A. Viola and M. J. Jones. Robust real-time face detection.
+	International Journal of Computer Vision
+	57(2):137–154, 2004.
 
-\newblock {\em Fluidos, Carlos Toro s.}
-
-\bibitem[http://www.windows2universe.org/]{Figueredo:2009dg}
-
-\newblock {\em Becca Hatheway}
-
-\bibitem[http://meteorologia.pucp.edu.pe/]{Figueredo:2009dg}
-
-\newblock {\em Hernan Castillo, Pedro Ríos}
-
-\bibitem[http://www.ammonit.com/Manuales]{Figueredo:2009dg}
-
-\newblock {\em Ammonit Measurement GmbH}
+	% the ICML 2013 workshop’s facial expression recognition challenge
+	\bibitem{replearning}
+	I. J. Goodfellow et al.
+	Challenges in representation learning: A report on three machine
+	learning contests.
+	Neural Networks
+	, 64:59 – 63, 2015.
+	Special Issue on ‘Deep Learning of Representations’
 

### 2018-03-03 ckplus_to_csv.py
ckplus to fer2013 class mapping

@@ -1,0 +1,87 @@
+#!/usr/bin/env python
+
+"""This module reads the CK+ dataset emotion labels and the corresponding
+png images into a CSV file for easier processing in higher level apis.
+All images are converted to 48x48 pixels grayscale.
+
+To run this module, you should have the same folder structure as shown in 
+README.md.
+
+    CK+ classes        FER classes         
+    0=neutral     =>   6
+    1=anger       =>   0  
+    2=contempt    =>   maybe angry or disgust?
+    3=disgust     =>   1
+    4=fear        =>   2
+    5=happy       =>   3
+    6=sadness     =>   4
+    7=surprise    =>   5
+
+FER labels are ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
+"""
+
+import os
+import glob
+import cv2
+
+
+def map_class(ck_class):
+    if ck_class == 0:
+        return 6
+    elif ck_class == 1:
+        return 0
+    elif ck_class == 2:
+        return 0
+    elif ck_class == 3:
+        return 1
+    elif ck_class == 4:
+        return 2
+    elif ck_class == 5:
+        return 3
+    elif ck_class == 6:
+        return 4
+    elif ck_class == 7:
+        return 5
+
+# csv file to write to
+csv_file = open('data/ckplus.csv', 'w')
+
+for root, dirs, files in os.walk("data/ckplus/labels"):
+    for file in files:
+        csv_line = ''
+        label_path = os.path.join(root, file)
+        # read labels
+        with open(label_path, 'r') as f:
+            label = map_class(int(float(f.read())))
+            csv_line += str(label) + ','
+            # print(label)
+        # get images for label
+        image_paths = root.replace('labels', 'images')
+        image_paths = sorted(glob.glob(image_paths + '/*.png'))
+        # last image is the peak expression
+        peak_image = cv2.imread(image_paths[-1], flags=cv2.IMREAD_COLOR)
+        neutral_image = cv2.imread(image_paths[0], flags=cv2.IMREAD_COLOR)
+
+        # convert images to grayscale
+        peak_image = cv2.cvtColor(peak_image, cv2.COLOR_BGR2GRAY)
+
+        # load OpenCV pre-trained cascade classifier for face detection
+        face_cascade = cv2.CascadeClassifier('model/haarcascade_frontalface_default.xml')
+
+        # find and extract faces on image (should be one face per image)
[…]

### 2018-03-03 model.py
parametrize img_size for convnet

@@ -1,15 +1,9 @@
 import tensorflow as tf
 
-
-# Image dimensions
-img_size = 48
-img_size_flat = img_size * img_size
-img_shape = (img_size, img_size)
-
 # Enable logging
 tf.logging.set_verbosity(tf.logging.INFO)
 
-def conv_net(x_dict, n_classes, dropout, reuse, is_training):
+def conv_net(x_dict, n_classes, img_size, dropout, reuse, is_training):
     
     # Define a scope for reusing the variables
     with tf.variable_scope('ConvNet', reuse=reuse):
@@ -67,8 +61,8 @@ def model_fn(features, labels, mode, params):
     # Build the neural network
     # Because Dropout have different behavior at training and prediction time, we
     # need to create 2 distinct computation graphs that still share the same weights.
-    logits_train = conv_net(features, params['num_classes'], params['dropout_rate'], reuse=False, is_training=True)
-    logits_test = conv_net(features, params['num_classes'], params['dropout_rate'], reuse=True, is_training=False)
+    logits_train = conv_net(features, params['num_classes'], params['img_size'], params['dropout_rate'], reuse=False, is_training=True)
+    logits_test = conv_net(features, params['num_classes'], params['img_size'], params['dropout_rate'], reuse=True, is_training=False)
     
     # Predictions
     pred_classes = tf.argmax(logits_test, axis=1)

### 2018-03-03 train.py
testing 128px img_size

@@ -50,6 +50,7 @@ estimator = tf.estimator.Estimator(
     params={
         'learning_rate': 0.001,
         'num_classes': 7,
+        'img_size': 48,
         'dropout_rate': 0.25
     })
 

### 2018-03-03 webcam.py

@@ -38,7 +38,7 @@ while True:
             print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
-            face_input = cv2.resize(face, (48, 48))
+            face_input = cv2.resize(face, (128, 128))
             face_input_byte = np.array(face_input, dtype=np.float32)
             face_input_intensity = np.divide(face_input_byte, 255)
 

### 2018-03-03 train_ckplus.py
train convnet on ckplus dataset

@@ -1,0 +1,64 @@
+#!/usr/bin/env python
+
+from sklearn.model_selection import train_test_split
+import tensorflow as tf
+import numpy as np
+import random
+import model
+import csv
+
+# Training Parameters
+num_steps = 250
+batch_size = 16
+
+# images and labels
+imgs = []
+labels = []
+
+# read csv file
+with open('data/ckplus.csv') as csvfile:
+    readCSV = csv.reader(csvfile, delimiter=',')
+    for row in readCSV:
+        pixels_test = [float(x) for x in row[1].split(' ')]
+        imgs.append(np.array(pixels_test))
+        labels.append(int(row[0]))
+
+# wrap into np arrays and divide pixels by 255
+imgs = np.array(imgs, dtype=np.float32)
+imgs = np.divide(imgs, 255)
+labels = np.array(labels, dtype=np.float32)
+
+# TODO: "leave one out" cross validation?
+
+# use sklearn for dividing dataset randomly (pseudo-random for now..)
+random.seed(42)
+imgs_train, imgs_test, labels_train, labels_test = train_test_split(imgs, labels, test_size=0.2, random_state=42)
+
+# Build the Estimator
+estimator = tf.estimator.Estimator(
+    model_fn=model.model_fn, 
+    model_dir='./model/ckplus',
+    params={
+        'learning_rate': 0.001,
+        'num_classes': 7,
+        'img_size': 128,
+        'dropout_rate': 0.25
+    })
+
+# Training
+# Define the input function for training
+input_fn = tf.estimator.inputs.numpy_input_fn(
+    x={'images': imgs_train}, y=labels_train,
+    batch_size=batch_size, num_epochs=None, shuffle=True)
+# Train the Model
+estimator.train(input_fn, steps=num_steps)
+
+
+
+# Evaluate the Model
+# Define the input function for evaluating
+input_fn = tf.estimator.inputs.numpy_input_fn(
+    x={'images': imgs_test}, y=labels_test,
+    batch_size=batch_size, shuffle=False)
+# Use the Estimator 'evaluate' method
+estimator.evaluate(input_fn)

### 2018-03-03 util/ckplus_to_csv.py
cleanup

@@ -1,70 +1,0 @@
-#!/usr/bin/env python
-
-"""This module reads the CK+ dataset emotion labels and the corresponding
-png images into a CSV file for easier processing in higher level apis.
-All images are converted to 48x48 pixels grayscale.
-
-To run this module, you should have the same folder structure as shown in 
-README.md.
-
-Labels are:
-    0=neutral
-    1=anger
-    2=contempt
-    3=disgust
-    4=fear
-    5=happy
-    6=sadness
-    7=surprise
-"""
-
-import os
-import glob
-import cv2
-
-# csv file to write to
-csv_file = open('../data/ckplus.csv', 'w')
-
-for root, dirs, files in os.walk("../data/ckplus/labels"):
-    for file in files:
-        csv_line = ''
-        label_path = os.path.join(root, file)
-        # read labels
-        with open(label_path, 'r') as f:
-            label = int(float(f.read()))
-            csv_line += str(label) + ','
-            # print(label)
-        # get images for label
-        image_paths = root.replace('labels', 'images')
-        image_paths = sorted(glob.glob(image_paths + '/*.png'))
-        # last image is the peak expression
-        peak_image = cv2.imread(image_paths[-1], flags=cv2.IMREAD_COLOR)
-        neutral_image = cv2.imread(image_paths[0], flags=cv2.IMREAD_COLOR)
-
-        # convert images to grayscale
-        peak_image = cv2.cvtColor(peak_image, cv2.COLOR_BGR2GRAY)
-        
-        if (label == 2):
-            cv2.imshow('Neutral Image', neutral_image)
-            cv2.imshow('Peak Image', peak_image)
-            cv2.waitKey()
-
-        # load OpenCV pre-trained cascade classifier for face detection
-        face_cascade = cv2.CascadeClassifier('../model/haarcascade_frontalface_default.xml')
-
-        # find and extract faces on image (should be one face per image)
-        # assuming all the falsely detected faces are smaller than the truly detected
-        # face, so discard all location params except of the largest
-        faces = face_cascade.detectMultiScale(peak_image, 1.1, 5)
-        (x, y, w, h) =  sorted(faces, key=len)[0]
-        # print(x, y, w, h)
-        face = peak_image[y:y+h, x:x+w]
-        # resize to 48x48 pixels
-        face = cv2.resize(face, (48, 48))
-
-        # write class and image pixels to csv
[…]

### 2018-03-05 ckplus_to_csv.py
neutral images

@@ -1,11 +1,14 @@
 #!/usr/bin/env python
 
 """This module reads the CK+ dataset emotion labels and the corresponding
-png images into a CSV file for easier processing in higher level apis.
+png images into a CSV file for easier processing in higher level apis. The
+first image in each sequence is the neutral expression while the last image
+is the peak expression.
+
 All images are converted to 48x48 pixels grayscale.
 
-To run this module, you should have the same folder structure as shown in 
-README.md.
+To run this module successfully, you should have the same folder 
+structure as shown in README.md.
 
     CK+ classes        FER classes         
     0=neutral     =>   6
@@ -23,7 +26,10 @@ FER labels are ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral
 import os
 import glob
 import cv2
+import random
 
+# load OpenCV pre-trained cascade classifier for face detection
+face_cascade = cv2.CascadeClassifier('model/haarcascade_frontalface_default.xml')
 
 def map_class(ck_class):
     if ck_class == 0:
@@ -42,6 +48,15 @@ def map_class(ck_class):
         return 4
     elif ck_class == 7:
         return 5
+
+def detect_face(img):
+    # find and extract faces on image (should be one face per image)
+    # assuming all the falsely detected faces are smaller than the actual
+    # face, so discard all location params except of the largest
+    faces = face_cascade.detectMultiScale(img, 1.1, 5)
+    (x, y, w, h) =  sorted(faces, key=len)[0]
+    # print(x, y, w, h)
+    return img[y:y+h, x:x+w]
 
 # csv file to write to
 csv_file = open('data/ckplus.csv', 'w')
@@ -64,24 +79,33 @@ for root, dirs, files in os.walk("data/ckplus/labels"):
 
         # convert images to grayscale
         peak_image = cv2.cvtColor(peak_image, cv2.COLOR_BGR2GRAY)
+        neutral_image = cv2.cvtColor(neutral_image, cv2.COLOR_BGR2GRAY)
 
-        # load OpenCV pre-trained cascade classifier for face detection
-        face_cascade = cv2.CascadeClassifier('model/haarcascade_frontalface_default.xml')
+        # detect faces
+        peak_face = detect_face(peak_image)
+        neutral_face = detect_face(neutral_image)
 
-        # find and extract faces on image (should be one face per image)
-        # assuming all the falsely detected faces are smaller than the truly detected
[…]

### 2018-03-05 train_ckplus.py
more steps

@@ -4,11 +4,12 @@ from sklearn.model_selection import train_test_split
 import tensorflow as tf
 import numpy as np
 import random
+import cv2
 import model
 import csv
 
 # Training Parameters
-num_steps = 250
+num_steps = 500
 batch_size = 16
 
 # images and labels
@@ -19,13 +20,19 @@ labels = []
 with open('data/ckplus.csv') as csvfile:
     readCSV = csv.reader(csvfile, delimiter=',')
     for row in readCSV:
-        pixels_test = [float(x) for x in row[1].split(' ')]
-        imgs.append(np.array(pixels_test))
+        pixels = [float(x) for x in row[1].split(' ')]
+        imgs.append(np.array(pixels))
         labels.append(int(row[0]))
 
 # wrap into np arrays and divide pixels by 255
 imgs = np.array(imgs, dtype=np.float32)
 imgs = np.divide(imgs, 255)
+
+# take a peek at the images
+# for i in imgs:
+#     cv2.imshow('Face', np.reshape(np.array(i), [128, 128]))
+#     cv2.waitKey()
+
 labels = np.array(labels, dtype=np.float32)
 
 # TODO: "leave one out" cross validation?

### 2018-03-10 webcam.py
shebang

@@ -1,3 +1,5 @@
+#!/usr/bin/env python
+
 import cv2
 import numpy as np
 import predict
@@ -35,7 +37,7 @@ while True:
         face_emotions = []
         for (x, y, w, h) in faces:
 
-            print(x, y, w, h)
+            # print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
             face_input = cv2.resize(face, (128, 128))

### 2018-03-10 model.py
saving the graph as a SavedModel for fast predictions in production

@@ -70,7 +70,9 @@ def model_fn(features, labels, mode, params):
     
     # If prediction mode, early return
     if mode == tf.estimator.ModeKeys.PREDICT:
-        return tf.estimator.EstimatorSpec(mode, predictions=pred_classes) 
+        return tf.estimator.EstimatorSpec(mode, 
+            predictions=pred_classes,
+            export_outputs={'classes': tf.estimator.export.PredictOutput(pred_classes)}) 
         
     # Define loss and optimizer
     loss_op = tf.reduce_mean(tf.nn.sparse_softmax_cross_entropy_with_logits(
@@ -93,5 +95,4 @@ def model_fn(features, labels, mode, params):
       loss=loss_op,
       train_op=train_op,
       eval_metric_ops={'accuracy': acc_op})
-
     return estim_specs

### 2018-03-10 predict.py

@@ -1,29 +1,12 @@
 import tensorflow as tf
+from tensorflow.contrib import predictor
 import numpy as np
 import model
 
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
-# Build the Estimator
-estimator = tf.estimator.Estimator(
-    model_fn=model.model_fn, 
-    model_dir='./model/ckplus',
-    params={
-        'learning_rate': 0.001,
-        'num_classes': 7,
-        'img_size': 128,
-        'dropout_rate': 0.25
-    })
-
-session = tf.Session()
+saved_model_predictor = predictor.from_saved_model(export_dir='saved_models/1520693695')
 
 def predictEmotion(img):
-    # Evaluate the Model
-    # Define the input function for evaluating
-    input_fn = tf.estimator.inputs.numpy_input_fn(
-        x={'images': img}, shuffle=False)
-
-    # Use the model to predict the images class
-    preds = list(estimator.predict(input_fn))
-    print(preds, " -> ", class_labels[preds[0]])
-    return preds[0]
+    output_dict = saved_model_predictor({'images': img})
+    return output_dict['output'][0]

### 2018-03-10 train_ckplus.py

@@ -9,7 +9,7 @@ import model
 import csv
 
 # Training Parameters
-num_steps = 500
+num_steps = 600
 batch_size = 16
 
 # images and labels
@@ -49,7 +49,7 @@ estimator = tf.estimator.Estimator(
         'learning_rate': 0.001,
         'num_classes': 7,
         'img_size': 128,
-        'dropout_rate': 0.25
+        'dropout_rate': 0.50
     })
 
 # Training
@@ -69,3 +69,12 @@ input_fn = tf.estimator.inputs.numpy_input_fn(
     batch_size=batch_size, shuffle=False)
 # Use the Estimator 'evaluate' method
 estimator.evaluate(input_fn)
+
+# Export the model as a SavedModel for production
+feature_spec = {'images': tf.placeholder(dtype=tf.float32, shape=[128, 128])}
+serving_input_fn = tf.estimator.export.build_raw_serving_input_receiver_fn(feature_spec)
+
+estimator.export_savedmodel(
+    export_dir_base='saved_models',
+    serving_input_receiver_fn=serving_input_fn,
+    as_text=True)

### 2018-03-11 model.py
remove todo

@@ -83,7 +83,6 @@ def model_fn(features, labels, mode, params):
     # Evaluate the accuracy of the model
     acc_op = tf.metrics.accuracy(labels=labels, predictions=pred_classes)
 
-    # TODO: Write summary for tensorboard
     tf.summary.scalar("loss", loss_op)
     tf.summary.scalar("accuracy", acc_op[0])
     

### 2018-03-11 predict.py
export savedmodel from estimator

@@ -5,7 +5,7 @@ import model
 
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
-saved_model_predictor = predictor.from_saved_model(export_dir='saved_models/1520693695')
+saved_model_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520799081')
 
 def predictEmotion(img):
     output_dict = saved_model_predictor({'images': img})

### 2018-03-11 train.py

@@ -1,3 +1,5 @@
+#!/usr/bin/env python
+
 import numpy as np
 import tensorflow as tf
 import csv
@@ -5,7 +7,7 @@ import csv
 import model
 
 # Training Parameters
-num_steps = 20000
+num_steps = 100
 batch_size = 128
 
 # Feature Vectors (0-255)
@@ -27,7 +29,7 @@ with open('data/fer2013.csv') as csvfile:
             pixels_train = [float(x) for x in row[1].split(' ')]
             imgs_train_byte.append(np.array(pixels_train))
             labels_train_class.append(int(row[0]))
-        elif row[2] == 'PrivateTest':
+        elif row[2] == 'PublicTest':
             pixels_test = [float(x) for x in row[1].split(' ')]
             imgs_test_byte.append(np.array(pixels_test))
             labels_test_class.append(int(row[0]))
@@ -71,3 +73,12 @@ input_fn = tf.estimator.inputs.numpy_input_fn(
     batch_size=batch_size, shuffle=False)
 # Use the Estimator 'evaluate' method
 estimator.evaluate(input_fn)
+
+# Export the model as a SavedModel for production use
+feature_spec = {'images': tf.placeholder(dtype=tf.float32, shape=[None, 48 * 48])}
+serving_input_fn = tf.estimator.export.build_raw_serving_input_receiver_fn(feature_spec)
+
+estimator.export_savedmodel(
+    export_dir_base='saved_models/fer2013',
+    serving_input_receiver_fn=serving_input_fn,
+    as_text=True)

### 2018-03-11 train_ckplus.py

@@ -9,12 +9,13 @@ import model
 import csv
 
 # Training Parameters
-num_steps = 600
-batch_size = 16
+num_steps = 1000
+batch_size = 64
 
 # images and labels
 imgs = []
 labels = []
+img_size = 128
 
 # read csv file
 with open('data/ckplus.csv') as csvfile:
@@ -30,7 +31,7 @@ imgs = np.divide(imgs, 255)
 
 # take a peek at the images
 # for i in imgs:
-#     cv2.imshow('Face', np.reshape(np.array(i), [128, 128]))
+#     cv2.imshow('Face', np.reshape(np.array(i), [img_size, img_size]))
 #     cv2.waitKey()
 
 labels = np.array(labels, dtype=np.float32)
@@ -38,7 +39,6 @@ labels = np.array(labels, dtype=np.float32)
 # TODO: "leave one out" cross validation?
 
 # use sklearn for dividing dataset randomly (pseudo-random for now..)
-random.seed(42)
 imgs_train, imgs_test, labels_train, labels_test = train_test_split(imgs, labels, test_size=0.2, random_state=42)
 
 # Build the Estimator
@@ -48,7 +48,7 @@ estimator = tf.estimator.Estimator(
     params={
         'learning_rate': 0.001,
         'num_classes': 7,
-        'img_size': 128,
+        'img_size': img_size,
         'dropout_rate': 0.50
     })
 
@@ -70,11 +70,11 @@ input_fn = tf.estimator.inputs.numpy_input_fn(
 # Use the Estimator 'evaluate' method
 estimator.evaluate(input_fn)
 
-# Export the model as a SavedModel for production
-feature_spec = {'images': tf.placeholder(dtype=tf.float32, shape=[128, 128])}
+# Export the model as a SavedModel for production use
+feature_spec = {'images': tf.placeholder(dtype=tf.float32, shape=[None, img_size * img_size])}
 serving_input_fn = tf.estimator.export.build_raw_serving_input_receiver_fn(feature_spec)
 
 estimator.export_savedmodel(
-    export_dir_base='saved_models',
+    export_dir_base='saved_models/ckplus',
     serving_input_receiver_fn=serving_input_fn,
     as_text=True)

### 2018-03-11 webcam.py

@@ -40,11 +40,14 @@ while True:
             # print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
-            face_input = cv2.resize(face, (128, 128))
+            face_input = cv2.resize(face, (128, 128)).reshape(1, 128*128)
             face_input_byte = np.array(face_input, dtype=np.float32)
             face_input_intensity = np.divide(face_input_byte, 255)
 
+            # pred_probas = predict.predictEmotion(face_input_intensity)
+            # np.set_printoptions(precision=3, suppress=True)
             pred_class = predict.predictEmotion(face_input_intensity)
+            print(pred_class)
             face_emotions.append(predict.class_labels[pred_class])
 
 

### 2018-03-12 predict.py
fer2013 saved model

@@ -5,8 +5,10 @@ import model
 
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
-saved_model_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520799081')
+fer2013_predictor = predictor.from_saved_model(export_dir='saved_models/fer2013/1520859040')
+# ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520799081')
 
 def predictEmotion(img):
-    output_dict = saved_model_predictor({'images': img})
-    return output_dict['output'][0]
+    output_dict_fer2013 = fer2013_predictor({'images': img})
+    # output_dict_ckplus = ckplus_predictor({'images': img})
+    return output_dict_fer2013['output'][0]

### 2018-03-12 train.py

@@ -7,7 +7,7 @@ import csv
 import model
 
 # Training Parameters
-num_steps = 100
+num_steps = 50000
 batch_size = 128
 
 # Feature Vectors (0-255)

### 2018-03-12 webcam.py

@@ -16,6 +16,8 @@ face_emotions = []
 frame_count = 0
 process_every_n_frames = 10
 
+img_size = 48
+
 while True:
     # Grab a single frame of the video
     ret, frame = webcam.read()
@@ -40,7 +42,7 @@ while True:
             # print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
-            face_input = cv2.resize(face, (128, 128)).reshape(1, 128*128)
+            face_input = cv2.resize(face, (img_size, img_size)).reshape(1, img_size*img_size)
             face_input_byte = np.array(face_input, dtype=np.float32)
             face_input_intensity = np.divide(face_input_byte, 255)
 

### 2018-03-12 train_ckplus.py
less steps

@@ -9,8 +9,8 @@ import model
 import csv
 
 # Training Parameters
-num_steps = 1000
-batch_size = 64
+num_steps = 500
+batch_size = 32
 
 # images and labels
 imgs = []

### 2018-03-12 predict.py
predictions from both models

@@ -6,7 +6,7 @@ import model
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
 fer2013_predictor = predictor.from_saved_model(export_dir='saved_models/fer2013/1520859040')
-# ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520799081')
+# ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520882345')
 
 def predictEmotion(img):
     output_dict_fer2013 = fer2013_predictor({'images': img})

### 2018-03-12 doc/paper.tex
thoughts on ferplus

@@
 Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI) and artificial emotional intelligence.
-The projects goal is to build a device to process live webcam images, detect faces and classify the emotion of each face.
+This projects goal is to develop a classifier for real-time detection of facial expressions.
 
@@
 As this has been done automatically, some images are not centered correctly.
-Faces are snapped from very different angles, hence generalization of these samples is difficult.
+Faces are depicted from very different angles, hence generalization of these samples is difficult.
 
@@
 
+The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivised, so we can expect label-noise to be significant.
+It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise).
+A research group at Microsoft (who?) decided to minimize label-noise in FER-2013 by re-labeling the dataset (REF).
+Ten workers on MTurk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.
+
 \subsection{Cohn-Kanade (CK+)}
-To argument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
+To augument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
 The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects.
 Each image sequence goes from a neutral facial expression to a "peak expression" according to the six basic emotions.
-Because only 327 of the 593 image sequences are labeled with emotion data.
+Only 327 of the 593 image sequences are labeled with emotion data.
 This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS) according to Ekman et al.
-CK+ also contains FACS data, in which action units (e.g raised eyebrow) are identified along with their degree of activation.
+CK+ also contains FACS data, in which facial action units (e.g raised eyebrow) are identified along with their degree of activation.
 
 Figure of CK+ distribution with neutral images.
+
+%------------------------------------------------
+
+\section{Related work}
 

### 2018-03-12 predict.py
fer and ckplus preds

@@ -6,9 +6,11 @@ import model
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
 fer2013_predictor = predictor.from_saved_model(export_dir='saved_models/fer2013/1520859040')
-# ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520882345')
+ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520882345')
 
-def predictEmotion(img):
-    output_dict_fer2013 = fer2013_predictor({'images': img})
-    # output_dict_ckplus = ckplus_predictor({'images': img})
+def predictEmotion(img_128, img_48):
+    output_dict_fer2013 = fer2013_predictor({'images': img_48})
+    output_dict_ckplus = ckplus_predictor({'images': img_128})
+    print("CK+ = ", output_dict_ckplus['output'][0])
+    print("FER = ", output_dict_fer2013['output'][0])
     return output_dict_fer2013['output'][0]

### 2018-03-12 webcam.py

@@ -16,8 +16,6 @@ face_emotions = []
 frame_count = 0
 process_every_n_frames = 10
 
-img_size = 48
-
 while True:
     # Grab a single frame of the video
     ret, frame = webcam.read()
@@ -42,13 +40,17 @@ while True:
             # print(x, y, w, h)
 
             face = frame_resized_gray[y:y+h, x:x+w]
-            face_input = cv2.resize(face, (img_size, img_size)).reshape(1, img_size*img_size)
-            face_input_byte = np.array(face_input, dtype=np.float32)
-            face_input_intensity = np.divide(face_input_byte, 255)
+            face_input_48 = cv2.resize(face, (48, 48)).reshape(1, 48*48)
+            face_input_48 = np.array(face_input_48, dtype=np.float32)
+            face_input_48 = np.divide(face_input_48, 255)
+
+            face_input_128 = cv2.resize(face, (128, 128)).reshape(1, 128*128)
+            face_input_128 = np.array(face_input_128, dtype=np.float32)
+            face_input_128 = np.divide(face_input_128, 255)
 
             # pred_probas = predict.predictEmotion(face_input_intensity)
             # np.set_printoptions(precision=3, suppress=True)
-            pred_class = predict.predictEmotion(face_input_intensity)
+            pred_class = predict.predictEmotion(face_input_128, face_input_48)
             print(pred_class)
             face_emotions.append(predict.class_labels[pred_class])
 

### 2018-04-01 doc/paper.tex
thoughts

@@
 Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI) and artificial emotional intelligence.
-This projects goal is to develop a classifier for real-time detection of facial expressions.
-
-As Convolutional Neural Networks (CNNs) have proven to be successful with many kinds of image classification  problems, I will focus on building a classifier using a state of the art CNN implementation.
+This projects goal is to develop a classifier for real-time detection of facial expressions using Convolutional Neural Networks (CNNs).
 
@@
 
-To build a classifier for facial expressions, a dataset with labeled images of faces is required.
+To build an automatic classifier for facial expressions, two datasets with labeled images of faces are required.
 Each of the datasets is split into two subsets: A training set that is used to fit the prediction and a test set to evaluate the prediction results.
@@
 
-The FER-2013 Dataset consists of roughly 37.000 images showing faces of people expressing a certain emotion.
+The FER-2013 dataset consists of roughly 37.000 images showing faces of people expressing a certain emotion.
 Most of the 48x48 pixel grayscale images are stock footage.
@@
 
-The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivised, so we can expect label-noise to be significant.
+The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so we can expect noisy labels to be a problem.
 It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise).
 A research group at Microsoft (who?) decided to minimize label-noise in FER-2013 by re-labeling the dataset (REF).
-Ten workers on MTurk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.
+Ten workers on Mechanical Turk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.
 
@@
 CK+ also contains FACS data, in which facial action units (e.g raised eyebrow) are identified along with their degree of activation.
+To show that extracted features are not strictly required for training an automatic classifier, the FACS data is not used as input.
 
 Figure of CK+ distribution with neutral images.
+
+%------------------------------------------------
+
+\section{Convolutional Neural Networks (CNNs)}
+Convolutional Neural Networks are are a special kind of multi-layer artificial neural networks.
+CNNs have been around since the 1990s, however they have lead to breakthrough results in practical small-scale application with computational power becoming more affordable in the last years.
+Convolutional Neural Networks can recognize visual patterns directly from pixel images with very little preprocessing.
+
+Like conventional Artificial Neural Networks, a CNN can be thought of as a function $f(x) = y$ where the input $x$ is an array of image pixel intensities and $y$ is a probability distribution of the classes this image is predicted to belong to (e.g happiness, surprise, sadness).
+
+\subsection{CNN Components}
+
+%------------------------------------------------
+
+\section{Model Architecture}
+A deep CNN is implemented as a prediction model for classifying images of faces.
+To keep the model architecture as simple as possible, advanced regularization techniques such as batch normalization have been omitted.
+The model architecture has been improved incrementally throughout the designing process.
+This has not been done programmatically but rather in a trial-and-error kind of way.
+
+\subsection{AlexNet-like CNN}
+For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.
 
@@
 \section{Discussion and Conclusion}
+
+% - CNN robustness to occlusion, distortions and geometric transformations.
 
@@
 
+	%[LeCun and Bengio, 1995a]
+    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
+    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995
+
 \end{thebibliography}

### 2018-04-09 util/ckplus_to_csv.py
csv header for pandas

@@ -63,6 +63,9 @@ def detect_face(img):
 # csv file to write to
 csv_file = open('data/ckplus.csv', 'w')
 
+# write csv header
+csv_file.write("class,pixels\n")
+
 for root, dirs, files in os.walk(abspath + '/../data/ckplus/labels'):
     for file in files:
         csv_line = ''

### 2018-04-09 doc/paper.tex
cnn components

@@
 
-Like conventional Artificial Neural Networks, a CNN can be thought of as a function $f(x) = y$ where the input $x$ is an array of image pixel intensities and $y$ is a probability distribution of the classes this image is predicted to belong to (e.g happiness, surprise, sadness).
+Like a regular Artificial Neural Network, a CNN can be thought of as a function $f(x) = y$ where the input $x$ is an image of a face and $y$ is a probability distribution of the classes this image is predicted to belong to (e.g happiness, surprise, sadness).
+Input images are flattened in dimension from 48x48 pixels to a single 2304-dimensional vector.
+For training and predictions, multiple images can be fed into the network function at once.
+When feeding $n$ images to the computational graph, the shape of the input tensor (matrix) is [2304, $n$].
+The associated output tensor is shaped [$n$, 7], where the second dimension is the probability for each of the seven emotion classes.
+
+To effectively reduce dimensions between the input and output space and find a valid representation of the input image, a CNN is implemented using the TensorFlow API.
+In the layers of the computational graph, a sequence of operations is applied to the input image.
 
 \subsection{CNN Components}
+
+\subsubsection{Convolutional Layer}
+The convolutional layer is the core building block of a CNN.
+Each layer is initialized with a set of $n$ random two-dimensional filter matrices (eg 3x3 pixels) and a pixel-stride.
+When an image is put into the convolutional layer, each of the filters is moved across the height and width of the image, calculating a dot product between the filter matrix and the current section of the image.
+The resulting matrices of dot products are called feature maps.
+To introduce non-linearity to the operation, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise.
+The filters in the convolutional layers are trained using a backpropagation algorithm (REF).
+Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyebrows or mouths.
+
+\subsubsection{Pooling Layer}
+To progressively reduce dimensionality and computational cost, pooling layers can be found in the computational graph.
+Similar to the convolutional layer, filters are moved across the layer input, executing a pooling operation on each position.
+A common pooling operation is taking the max value on every filter position (Max-Pooling).
+For instance when pooling with a filter size of 3x3 pixels and a stride of 2 is applied to a 48x48 pixels image, the pooling output is a 24x24 pixels image.
+
+\subsubsection{Fully-Connected layers}
+In CNNs, fully connected layers are usually added at the top of the network.
+The advanced features that have been extracted in the top convolutional layers are mapped to a few layers of weighted nodes.
+On the first fully-connected layer, linear combinations of the feature maps activations are created.
+On top of that, linear transformations are applied to reduce the dimension to match the number of classes.
+A softmax function can be applied on the output layer to normalize the topmost fully connected layer into a valid probability distribution.
 

### 2018-04-11 model.py
prediction probability output, remove argmax

@@ -71,7 +71,7 @@ def model_fn(features, labels, mode, params):
     if mode == tf.estimator.ModeKeys.PREDICT:
         spec = tf.estimator.EstimatorSpec(mode, 
             predictions=pred_classes,
-            export_outputs={'classes': tf.estimator.export.PredictOutput(pred_classes)})
+            export_outputs={'classes': tf.estimator.export.PredictOutput(pred_probas)})
     else:
         # Define loss across all classes
         loss_op = tf.reduce_mean(

### 2018-04-11 predict.py

@@ -5,15 +5,21 @@ import model
 
 class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
 
-fer2013_predictor = predictor.from_saved_model(export_dir='saved_models/fer2013/1520859040')
-ferplus_predictor = predictor.from_saved_model(export_dir='saved_models/ferplus/1521543204')
-ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520882345')
+# fer2013_predictor = predictor.from_saved_model(export_dir='saved_models/fer2013/1520859040')
+ferplus_predictor = predictor.from_saved_model(export_dir='saved_models/ferplus/1523392049')
+# ckplus_predictor = predictor.from_saved_model(export_dir='saved_models/ckplus/1520882345')
 
 def predictEmotion(img_128, img_48):
-    output_dict_fer2013 = fer2013_predictor({'images': img_48})
+    # output_dict_fer2013 = fer2013_predictor({'images': img_48})
     output_dict_ferplus = ferplus_predictor({'images': img_48})
-    output_dict_ckplus = ckplus_predictor({'images': img_128})
-    print("CK+ = ", output_dict_ckplus['output'][0])
-    print("FER = ", output_dict_fer2013['output'][0])
+    # output_dict_ckplus = ckplus_predictor({'images': img_128})
+    # print("CK+ = ", output_dict_ckplus['output'][0])
+    # print("FER = ", output_dict_fer2013['output'][0])
     print("FER+ = ", output_dict_ferplus['output'][0])
+    print(output_dict_ferplus)
     return output_dict_ferplus['output'][0]
+
+def predict_emotion_probas(img_128, img_48, n=3):
+    output_dict_ferplus = ferplus_predictor({'images': img_48})
+    class_indices = np.argsort(output_dict_ferplus['output'][0])
+    return (class_indices[::-1], output_dict_ferplus['output'][0])

### 2018-04-11 webcam.py
top 3 emotions and bar visualization for probas

@@ -12,16 +12,23 @@ face_cascade = cv2.CascadeClassifier('model/haarcascade_frontalface_default.xml'
 
 # variables
 face_locations = []
-face_emotions = []
+faces_preds = []
 frame_count = 0
-process_every_n_frames = 10
+process_every_n_frames = 20
+
+# colors
+white = (255, 255, 255)
+black = (0, 0, 0)
+class_labels = ['Angry', 'Disgust', 'Fear', 'Happy', 'Sad', 'Surprise', 'Neutral']
+emotion_colors = [(54, 67, 244), (136, 150, 0), (76, 39, 156), (80, 175, 76), (0, 152, 255), (59, 235, 255), (243, 150, 33)]
+
 
 while True:
     # Grab a single frame of the video
     ret, frame = webcam.read()
 
     # Resize frame of video to 1/4 size for faster face recognition processing
-    frame_resized = cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
+    frame_resized = frame # cv2.resize(frame, (0, 0), fx=0.25, fy=0.25)
 
     # Only process every other frame of video to save time
     if frame_count % process_every_n_frames == 0:
@@ -34,7 +41,7 @@ while True:
 
         print('Faces found: ', len(faces))
 
-        face_emotions = []
+        faces_preds = []
         for (x, y, w, h) in faces:
 
             # print(x, y, w, h)
@@ -50,26 +57,41 @@ while True:
 
             # pred_probas = predict.predictEmotion(face_input_intensity)
             # np.set_printoptions(precision=3, suppress=True)
-            pred_class = predict.predictEmotion(face_input_128, face_input_48)
-            print(pred_class)
-            face_emotions.append(predict.class_labels[pred_class])
+            prediction = predict.predict_emotion_probas(face_input_128, face_input_48)
+            print(prediction)
+            faces_preds.append(prediction)
 
 
     # Display the results
-    for (x, y, w, h), name in zip(faces, face_emotions):
+    for (x, y, w, h), (classes, probas) in zip(faces, faces_preds):
 
-        x *= 4
-        y *= 4
-        w*= 4
-        h *= 4
+        # x *= 4
+        # y *= 4
+        # w *= 4
+        # h *= 4
 
         # Draw a box around the face
[…]

### 2018-04-12 doc/paper.tex
re-structuring sections

@@
 
-Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction (HCI) and artificial emotional intelligence.
-This projects goal is to develop a classifier for real-time detection of facial expressions using Convolutional Neural Networks (CNNs).
+Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction and artificial emotional intelligence.
+Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise [REF].
+This projects goal is to develop a classifier for real-time detection of these facial expressions.
+
+%------------------------------------------------
+
+\section{Related work}
+In the last years, great progress was made on the problem of automatically detecting facial expression from images.
+Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation [REF].
+SIFT aggregation [REF]… Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data.
+This can be accomplished with Convolutional Neural Networks (CNNs) that are trained on a set of labeled images.
 
@@
 \section{Datasets}
-
-Paul Ekmans 1969 research paper "Facial Expressions" (REF) elaborates that facial expressions can be interpreted and categorized into six classes of basic emotion that can be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise.
-In this project, these classes are used for training and predictions.
-
-To build an automatic classifier for facial expressions, two datasets with labeled images of faces are required.
-Each of the datasets is split into two subsets: A training set that is used to fit the prediction and a test set to evaluate the prediction results.
+To train an automatic classifier using deep learning technology, a dataset that describes the problem is required.
+Each image in the dataset is mapped to a matching integer class label for the corresponding emotion.
+Two datasets, FER2013 [REF] and CK+ [REF] have been considered for training the deep learning model.
 
 \subsection{FER-2013}
-
 The FER-2013 dataset consists of roughly 37.000 images showing faces of people expressing a certain emotion.
@@
 \subsubsection{Fully-Connected layers}
-In CNNs, fully connected layers are usually added at the top of the network.
-The advanced features that have been extracted in the top convolutional layers are mapped to a few layers of weighted nodes.
+In CNNs, fully connected layers are added at the top of the network.
+The advanced features that have been extracted in the top convolutional layers are usually connected to two or more layers of weighted nodes.
 On the first fully-connected layer, linear combinations of the feature maps activations are created.
@@
 
-\section{Model Architecture}
+\section{Model Architecture and Training}
 A deep CNN is implemented as a prediction model for classifying images of faces.
-To keep the model architecture as simple as possible, advanced regularization techniques such as batch normalization have been omitted.
 The model architecture has been improved incrementally throughout the designing process.
-This has not been done programmatically but rather in a trial-and-error kind of way.
+Hyperparameter Optimization has not been done programmatically but rather in a trial-and-error kind of way.
 
@@
 
-%------------------------------------------------
-
-\section{Related work}
+\subsection{Training Procedure}
 
@@
 
+    % Paul Ekmans 1969 research paper "Facial Expressions"
+
 \end{thebibliography}

### 2018-04-12 doc/paper.tex
references and discussion

@@
 \usepackage{hyperref} % For hyperlinks in the PDF
+\usepackage{cite} % BibTex
 
@@
 Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction and artificial emotional intelligence.
-Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise [REF].
+Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}.
 This projects goal is to develop a classifier for real-time detection of these facial expressions.
@@
 In the last years, great progress was made on the problem of automatically detecting facial expression from images.
-Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation [REF].
-SIFT aggregation [REF]… Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data.
+Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation \cite{lien98}.
+Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data.
 This can be accomplished with Convolutional Neural Networks (CNNs) that are trained on a set of labeled images.
@@
 Each image in the dataset is mapped to a matching integer class label for the corresponding emotion.
-Two datasets, FER2013 [REF] and CK+ [REF] have been considered for training the deep learning model.
+Two datasets, FER2013 \cite{goodfel13} and CK+ \cite{cohn00} have been considered for training the deep learning model.
 
@@
 It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise).
-A research group at Microsoft (who?) decided to minimize label-noise in FER-2013 by re-labeling the dataset (REF).
+A research group at Microsoft decided to minimize label-noise in FER-2013 by re-labeling the dataset \cite{barsoum16}.
 Ten workers on Mechanical Turk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.
@@
 \subsection{Cohn-Kanade (CK+)}
-To augument the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
+To augment the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
 The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects.
@@
 To introduce non-linearity to the operation, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise.
-The filters in the convolutional layers are trained using a backpropagation algorithm (REF).
+The filters in the convolutional layers are trained using a backpropagation algorithm.
 Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyebrows or mouths.
@@
 
-\subsection{AlexNet-like CNN}
-For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.
+%\subsection{AlexNet-like CNN}
+%For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.
 
@@
 % - CNN robustness to occlusion, distortions and geometric transformations.
+The final model scored an accuracy of ~75\% on a test-set of 3557 samples.
+In live predictions with a webcam, the classifier has shown to be a cool gadget to play around with.
+However, some emotions such as disgust and fear are hard to reproduce predictions for.
+This is most likely because the seven classes of emotion in the dataset are highly imbalanced.
+For instance disgust only has 500 samples while happy has much more.
+This results in biased predictions with higher probabilities of the majority classes.
+In order to balance the training dataset, data augmentation could be used: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
+For normalizing the images, it is a common practice to subtract the mean pixel intensities over the whole dataset from each picture.
+Given more time for this project, I would like to have experimented with such techniques of data augmentation.
 
@@
 
-\begin{thebibliography}{9}
+%\begin{thebibliography}{9}
+%
+%	% OpenCV Haarcascade Classifier
+%	\bibitem{violajones}
+%  	P. A. Viola and M. J. Jones. Robust real-time face detection.
+%	International Journal of Computer Vision
+%	57(2):137–154, 2004.
+%
+%	% the ICML 2013 workshop’s facial expression recognition challenge
+%	\bibitem{replearning}
+%	I. J. Goodfellow et al.
+%	Challenges in representation learning: A report on three machine
+%	learning contests.
+%	Neural Networks
+%	, 64:59 – 63, 2015.
+%	Special Issue on ‘Deep Learning of Representations’
+%	
+%	%[LeCun and Bengio, 1995a]
+%    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
+%    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995
+%    
+%    % Paul Ekmans 1969 research paper "Facial Expressions"
+%
+% 
+%\end{thebibliography}
 
-	% OpenCV Haarcascade Classifier
-	\bibitem{violajones}
-  	P. A. Viola and M. J. Jones. Robust real-time face detection.
-	International Journal of Computer Vision
-	57(2):137–154, 2004.
-
-	% the ICML 2013 workshop’s facial expression recognition challenge
-	\bibitem{replearning}
-	I. J. Goodfellow et al.
-	Challenges in representation learning: A report on three machine
-	learning contests.
-	Neural Networks
-	, 64:59 – 63, 2015.
-	Special Issue on ‘Deep Learning of Representations’
-
-	%[LeCun and Bengio, 1995a]
-    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
-    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995
-
-    % Paul Ekmans 1969 research paper "Facial Expressions"
-
-\end{thebibliography}
+\bibliography{paper}
+\bibliographystyle{plain}
 

### 2018-04-14 doc/paper.tex
figures

@@
 
+\begin{figure}[H]
+	\includegraphics[width=0.48\textwidth]{fer_examples}
+	\caption{FER-2013 training samples}
+\end{figure}
+
 \subsection{FER-2013}
@@
 \begin{figure}[H]
-	\includegraphics[width=0.48\textwidth]{class_dist}
-	\caption{FER-2013 emotion class distribution}
+	\includegraphics[width=0.48\textwidth]{ferplus_distribution}
+	\caption{FER-2013 dataset class distribution}
 \end{figure}
@@
 
-Figure of CK+ distribution with neutral images.
+\begin{figure}[H]
+	\includegraphics[width=0.48\textwidth]{ckplus_examples}
+	\caption{Posed facial expressions from the CK+ dataset}
+\end{figure}
 
@@
 The filters in the convolutional layers are trained using a backpropagation algorithm.
-Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyebrows or mouths.
+Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyes or mouths.
 

### 2018-04-15 doc/paper.tex
corrections

@@
 
-Automatically recognizing facial expressions is an interesting and challenging problem in many fields, including human computer interaction and artificial emotional intelligence.
+Automatically recognizing facial expressions is an interesting and challenging problem in the fields human computer interaction and artificial emotional intelligence.
 Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}.
@@
 Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation \cite{lien98}.
-Manual feature extraction tends to require expert knowledge and extra work for implementing complex algorithms, so it is convenient to implicitly extract relevant features from the data.
-This can be accomplished with Convolutional Neural Networks (CNNs) that are trained on a set of labeled images.
-
+Manually extracting the relevant features from the training data requires expert knowledge or complex algorithms.
+In Convolutional Neural Networks (CNNs), features can be learned from the training data implicitly and only the image pixels are required as input for predictions.
+CNNs and traditional Support Vector Machines have been used for various image recognition tasks, including facial expression recognition \cite{dumas01}.
 %------------------------------------------------
@@
 \section{Datasets}
-To train an automatic classifier using deep learning technology, a dataset that describes the problem is required.
-Each image in the dataset is mapped to a matching integer class label for the corresponding emotion.
-Two datasets, FER2013 \cite{goodfel13} and CK+ \cite{cohn00} have been considered for training the deep learning model.
+To train an automatic classifier using deep learning technology, a dataset that approximates the problem is required.
+Each image in the dataset is mapped to an integer class label for the corresponding emotion.
+Two datasets, FER2013 \cite{goodfel13} and CK+ \cite{cohn00} are considered for training the deep learning model.
 
@@
 Most of the 48x48 pixel grayscale images are stock footage.
-These images have been centered on the persons faces to show a similar region in every sample.
+All images have been centered on the persons faces to show a similar region in every sample.
 As this has been done automatically, some images are not centered correctly.
-Faces are depicted from very different angles, hence generalization of these samples is difficult.
+Persons are depicted from very different camera angles, some faces are occluded and rotation is different across the images.
+These problems make generalization of the training data difficult.
 
@@
 
-The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so we can expect noisy labels to be a problem.
-It is expected that every subject shows only a single emotion, where in reality faces can express mixed emotions (e.g happiness and surprise).
-A research group at Microsoft decided to minimize label-noise in FER-2013 by re-labeling the dataset \cite{barsoum16}.
-Ten workers on Mechanical Turk were asked to label each image, yielding a probability distribution of emotion classes instead of a single class for each image.
+The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so faulty labels are likely a problem.
+It is expected that every subject shows only a single emotion, where in reality, faces can express at least two emtions (e.g happiness and surprise) simultaniously.
+To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft worked on re-labeling the dataset \cite{barsoum16}: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
+Instead of single class ground-truths, probability distributions of emotion classes describe each persons facial expression.
 
 \subsection{Cohn-Kanade (CK+)}
-To augment the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is considered.
+To augment the wild variety of FER-2013 images, a second, more conservative dataset with only frontal faces is used to train a second classifier.
 The Cohn-Kanade (CK+) dataset consists of 593 image sequences across 123 subjects.
-Each image sequence goes from a neutral facial expression to a "peak expression" according to the six basic emotions.
+Each image sequence goes from a neutral facial expression to a posed "peak expression" according to the six basic emotions.
 Only 327 of the 593 image sequences are labeled with emotion data.
-This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS) according to Ekman et al.
-CK+ also contains FACS data, in which facial action units (e.g raised eyebrow) are identified along with their degree of activation.
+This is because not all images fit the prototypic definition of the Facial Action Coding System (FACS).
+CK+ also comes with FACS data, where facial action units (e.g raised eyebrow) are identified along with their degree of activation.
 To show that extracted features are not strictly required for training an automatic classifier, the FACS data is not used as input.
@@
 Convolutional Neural Networks are are a special kind of multi-layer artificial neural networks.
-CNNs have been around since the 1990s, however they have lead to breakthrough results in practical small-scale application with computational power becoming more affordable in the last years.
+CNNs have been around since the 1990s \cite{lecun98}, however they have lead to breakthrough results in practical small-scale application with computational power becoming more affordable in the last years.
 Convolutional Neural Networks can recognize visual patterns directly from pixel images with very little preprocessing.
@@
 Input images are flattened in dimension from 48x48 pixels to a single 2304-dimensional vector.
-For training and predictions, multiple images can be fed into the network function at once.
-When feeding $n$ images to the computational graph, the shape of the input tensor (matrix) is [2304, $n$].
-The associated output tensor is shaped [$n$, 7], where the second dimension is the probability for each of the seven emotion classes.
+For training and predictions, multiple images can be fed into the neural network at once.
 
-To effectively reduce dimensions between the input and output space and find a valid representation of the input image, a CNN is implemented using the TensorFlow API.
-In the layers of the computational graph, a sequence of operations is applied to the input image.
+The model function is implemented as a multi-layer computational graph where transformations are applied to the input on every layer.
 
@@
 The filters in the convolutional layers are trained using a backpropagation algorithm.
-Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or, in the top convolutional layers, more complex patterns like eyes or mouths.
+Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or more complex patterns like eyes or mouths in the top convolutional layers.
 
@@
 On the first fully-connected layer, linear combinations of the feature maps activations are created.
-On top of that, linear transformations are applied to reduce the dimension to match the number of classes.
+On top of that, linear transformations are applied to reduceach layer transforms the output of the previouse the dimension to match the number of classes.
 A softmax function can be applied on the output layer to normalize the topmost fully connected layer into a valid probability distribution.
@@
 \section{Model Architecture and Training}
-A deep CNN is implemented as a prediction model for classifying images of faces.
-The model architecture has been improved incrementally throughout the designing process.
-Hyperparameter Optimization has not been done programmatically but rather in a trial-and-error kind of way.
+To effectively reduce dimensions between the input and output space and find a valid representation of the input image, a CNN is implemented using the TensorFlow higher-level APIs.
+The TensorFlow extimator API provides a convenient abstraction layer by encapsulating trainable classifiers.
+
+In simple CNNs, it is common to combine convolutional and pooling layers to compute feature activations while reducing dimensionality.
+During the designing process, it turned out that stacking multiple convolutional layers in a VGG-Fashion \cite{simonyan14} improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
+
+The model architecture has been improved incrementally throughout the designing process. Hyperparameter optimization has not been done programmatically but rather by trial-and-error.
 
@@
 \subsection{Training Procedure}
+The CNN is trained on a subset of the FER-dataset according to the original train/test/validation split with the FERplus prabability distributions as labels.
+The models filters and weights/biases of the fully connected layers are fit to the training set using a backpropagation algorithm with the Adam-optimizer minimizing a cross-entropy loss function.
+To prevent overfitting the model on the training set, the loss value on the validation set (unseen data) is computed periodically during training.
+When the validation loss stops decreasing, training is stopped and the final model accuracy is computed on the test subset.
 
@@
 The final model scored an accuracy of ~75\% on a test-set of 3557 samples.
-In live predictions with a webcam, the classifier has shown to be a cool gadget to play around with.
-However, some emotions such as disgust and fear are hard to reproduce predictions for.
+In live predictions with a webcam, the classifier has shown to be surprisingly accurate and is a cool gadget to play around with.
+However, predictions for the classes $disgust$ and $fear$ are hard to reproduce.
 This is most likely because the seven classes of emotion in the dataset are highly imbalanced.

### 2018-04-17 doc/paper.tex
face detection

@@
 This projects goal is to develop a classifier for real-time detection of these facial expressions.
+To simplify the process and focus on the main problem, faces are being detected using the OpenCV Haar feature-based cascade classifier \cite{violajones04}.
 
@@
 In simple CNNs, it is common to combine convolutional and pooling layers to compute feature activations while reducing dimensionality.
-During the designing process, it turned out that stacking multiple convolutional layers in a VGG-Fashion \cite{simonyan14} improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
+During the designing process, it turned out that stacking multiple convolutional layers in improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
 
@@
 %	% OpenCV Haarcascade Classifier
-%	\bibitem{violajones}
-%  	P. A. Viola and M. J. Jones. Robust real-time face detection.
-%	International Journal of Computer Vision
-%	57(2):137–154, 2004.
+
 %

### 2018-04-17 doc/paper.tex
reading through and correcting

--- 2018-04-15 doc/paper.tex
@@
 
-Automatically recognizing facial expressions is an interesting and challenging problem in the fields human computer interaction and artificial emotional intelligence.
+Automatically recognizing facial expressions is an interesting and challenging problem in the field of human computer interaction.
 Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}.
@@
 Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation \cite{lien98}.
-Manually extracting the relevant features from the training data requires expert knowledge or complex algorithms.
+However, manually extracting these relevant features from the training data requires expert knowledge or complex algorithms.
 In Convolutional Neural Networks (CNNs), features can be learned from the training data implicitly and only the image pixels are required as input for predictions.
-CNNs and traditional Support Vector Machines have been used for various image recognition tasks, including facial expression recognition \cite{dumas01}.
+Due to this convenience, CNNs and have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
 %------------------------------------------------
@@
 The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so faulty labels are likely a problem.
-It is expected that every subject shows only a single emotion, where in reality, faces can express at least two emtions (e.g happiness and surprise) simultaniously.
+It is expected that every subject shows only a single emotion, where in reality faces can express at least two emtions simultaniously (e.g happiness and surprise).
 To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft worked on re-labeling the dataset \cite{barsoum16}: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
-Instead of single class ground-truths, probability distributions of emotion classes describe each persons facial expression.
+Instead of single class labels, probability distributions of emotion classes describe each persons facial expression.
 
@@
 The resulting matrices of dot products are called feature maps.
-To introduce non-linearity to the operation, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise.
+To introduce non-linearity to the model function, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise.
 The filters in the convolutional layers are trained using a backpropagation algorithm.
@@
 \subsubsection{Pooling Layer}
-To progressively reduce dimensionality and computational cost, pooling layers can be found in the computational graph.
+To progressively reduce dimensionality and computational cost, pooling layers are introduced in the computational graph.
 Similar to the convolutional layer, filters are moved across the layer input, executing a pooling operation on each position.
@@
 On the first fully-connected layer, linear combinations of the feature maps activations are created.
-On top of that, linear transformations are applied to reduceach layer transforms the output of the previouse the dimension to match the number of classes.
+Intuitively, combinations of filters are mapped to emotion class probabilities.
 A softmax function can be applied on the output layer to normalize the topmost fully connected layer into a valid probability distribution.

### 2018-04-17 doc/paper.tex
remove vgg reference

@@
 In simple CNNs, it is common to combine convolutional and pooling layers to compute feature activations while reducing dimensionality.
-During the designing process, it turned out that stacking multiple convolutional layers in a VGG-Fashion \cite{simonyan14} improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
+During the designing process, it turned out that stacking multiple convolutional layers improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
 
@@
 
-\section{Discussion and Conclusion}
+\section{Conclusion and Discussion}
 

### 2018-04-30 doc/paper.tex
revisions

@@
 %----------------------------------------------------------------------------------------
-%	ABSTRACT
-%----------------------------------------------------------------------------------------
-
-\begin{abstract}
-
-\noindent El barómetro es un instrumento de medición atmosérica, específicamente utilizado en la determinación de la fuerza por unidad de superficie ejercida por el peso de la atmósfera.
-Existe un gran número de equipos atmoféricos con distintos tipos de estos aparátos y son diariamente utilizados ya que la presión atmosférica juega un papel importante en la determinación y pronóstico del tiempo así como en el área de investigación al momento de realizarse experimentos ya que pueden llegar a afectar o hacer variar el funcionamiento de muchos aparatos electrónicos y mecánicos.
-
-\end{abstract}
-
-%----------------------------------------------------------------------------------------
 %	ARTICLE CONTENTS
@@
 Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}.
-This projects goal is to develop a classifier for real-time detection of these facial expressions.
+This projects goal is to develop a classifier for real-time detection of these facial expressions using deep learning.
 
@@
 In Convolutional Neural Networks (CNNs), features can be learned from the training data implicitly and only the image pixels are required as input for predictions.
-Due to this convenience, CNNs and have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
+With this convenience, CNNs have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
 %------------------------------------------------
@@
 \section{Datasets}
-To train an automatic classifier using deep learning technology, a dataset that approximates the problem is required.
-Each image in the dataset is mapped to an integer class label for the corresponding emotion.
+To train an automatic classifier using deep learning technology, a dataset with samples that describe the problem is required.
+Each image in the dataset is mapped to an integer class label 0–6 for the corresponding emotion.
 Two datasets, FER2013 \cite{goodfel13} and CK+ \cite{cohn00} are considered for training the deep learning model.
@@
 As this has been done automatically, some images are not centered correctly.
-Persons are depicted from very different camera angles, some faces are occluded and rotation is different across the images.
-These problems make generalization of the training data difficult.
+Persons are depicted from different camera angles, some faces are occluded and rotation is different across the images.
+These circumstances make the problem of generalizing on the training data difficult.
 
@@
 CK+ also comes with FACS data, where facial action units (e.g raised eyebrow) are identified along with their degree of activation.
-To show that extracted features are not strictly required for training an automatic classifier, the FACS data is not used as input.
 
@@
 Convolutional Neural Networks are are a special kind of multi-layer artificial neural networks.
-CNNs have been around since the 1990s \cite{lecun98}, however they have lead to breakthrough results in practical small-scale application with computational power becoming more affordable in the last years.
+CNNs have been around since the 1990s \cite{lecun98}, however they have lead to breakthrough results in practical application with computational power becoming more affordable in the last years.
 Convolutional Neural Networks can recognize visual patterns directly from pixel images with very little preprocessing.
@@
 The resulting matrices of dot products are called feature maps.
-To introduce non-linearity to the model function, a ReLu activation function $f(n) = max(0, n)$ is used on the feature-maps element-wise.
-The filters in the convolutional layers are trained using a backpropagation algorithm.
+\cite{karpathy} To introduce non-linearity to the model function, a ReLu activation function $f(x) = max(0, x)$ is used on the feature-maps element-wise.
+The filters in the convolutional layers are trained using the backpropagation algorithm \cite{goodfel16}.
 Intuitively, filters activate when they see different types of visual features, such as vertical or diagonal edges on the lower layers or more complex patterns like eyes or mouths in the top convolutional layers.
@@
 \subsubsection{Pooling Layer}
-To progressively reduce dimensionality and computational cost, pooling layers are introduced in the computational graph.
+To down-sample images between the layers in favor of making the training process faster, pooling layers are introduced in the computational graph.
 Similar to the convolutional layer, filters are moved across the layer input, executing a pooling operation on each position.
 A common pooling operation is taking the max value on every filter position (Max-Pooling).
-For instance when pooling with a filter size of 3x3 pixels and a stride of 2 is applied to a 48x48 pixels image, the pooling output is a 24x24 pixels image.
+For instance when pooling with a filter size of 2x2 pixels and a stride of 2 is applied to a 48x48 pixels image, the pooling output is a 24x24 pixels image.
 
@@
 To effectively reduce dimensions between the input and output space and find a valid representation of the input image, a CNN is implemented using the TensorFlow higher-level APIs.
-The TensorFlow extimator API provides a convenient abstraction layer by encapsulating trainable classifiers.
+The TensorFlow extimator API is used, which provides a convenient abstraction layer by encapsulating trainable classifiers.
 
@@
 During the designing process, it turned out that stacking multiple convolutional layers improved the model accuracy on the test set by about 20 percent compared to alternating pooling and convolutional layers.
+This is because the pooling operation causes loss of information and multiple adjacent convolutional layers allow more complex filters to emerge.
 
@@
 
-%\subsection{AlexNet-like CNN}
-%For the Imagenet Challenge (REF) Alex … proposed a deep neural network (DNN) model consisting of alternating convolution and pooling layers with local response normalization in between, followed by three fully connected layers.
+\bigskip
+
+\begin{tabular}{|| c ||}
+
+  \hline
+  FC [7 nodes] output\\
+  FC [1024 nodes]\\
+  \hline
+  MAX POOL [2x2 pixels, stride 2]\\
+  CONV [256 filters, 3x3 pixels]\\
+  CONV [256 filters, 3x3 pixels]\\
+  CONV [256 filters, 3x3 pixels]\\
+  \hline
+  MAX POOL [2x2 pixels, stride 2]\\
+  CONV [128 filters, 3x3 pixels]\\
+  CONV [128 filters, 3x3 pixels]\\
+  CONV [128 filters, 3x3 pixels]\\
+  \hline
+  MAX POOL [2x2 pixels, stride 2]\\
+  CONV [64 filters, 3x3 pixels]\\
+  CONV [64 filters, 3x3 pixels]\\
+  \hline
+  MAX POOL [2x2 pixels, stride 2]\\
+  CONV [32 filters, 3x3 pixels]\\
+  CONV [32 filters, 3x3 pixels]\\
+  \hline
+
+\end{tabular}
+
+\medskip
+
+\textit{The final network architecture. Bottom-up from input to output.}
 
@@
 The CNN is trained on a subset of the FER-dataset according to the original train/test/validation split with the FERplus prabability distributions as labels.
-The models filters and weights/biases of the fully connected layers are fit to the training set using a backpropagation algorithm with the Adam-optimizer minimizing a cross-entropy loss function.
+The models filters and weights/biases of the fully connected layers are fit to the training set using a backpropagation algorithm.
+Therefore, a cross-entropy loss function is used to measure the model error and the Adam optimization algorithm with an initial step size of 0.001 is used to minimize the loss.
 To prevent overfitting the model on the training set, the loss value on the validation set (unseen data) is computed periodically during training.
@@
 The final model scored an accuracy of ~75\% on a test-set of 3557 samples.
-In live predictions with a webcam, the classifier has shown to be surprisingly accurate and is a cool gadget to play around with.
+In live predictions with a webcam, the classifier has shown to be accurate.
 However, predictions for the classes $disgust$ and $fear$ are hard to reproduce.
@@
 This results in biased predictions with higher probabilities of the majority classes.
-In order to balance the training dataset, data augmentation could be used: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
-For normalizing the images, it is a common practice to subtract the mean pixel intensities over the whole dataset from each picture.
+In order to balance the training dataset, data augmentation could be used \cite{krizhevsky12}: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
 Given more time for this project, I would like to have experimented with such techniques of data augmentation.
@@
 %----------------------------------------------------------------------------------------
-
-%\begin{thebibliography}{9}
-%
-%	% OpenCV Haarcascade Classifier
-%	\bibitem{violajones}
-%  	P. A. Viola and M. J. Jones. Robust real-time face detection.
-%	International Journal of Computer Vision
-%	57(2):137–154, 2004.
-%
-%	% the ICML 2013 workshop’s facial expression recognition challenge
-%	\bibitem{replearning}
-%	I. J. Goodfellow et al.
-%	Challenges in representation learning: A report on three machine
-%	learning contests.
-%	Neural Networks
-%	, 64:59 – 63, 2015.
-%	Special Issue on ‘Deep Learning of Representations’
-%	
-%	%[LeCun and Bengio, 1995a]
-%    %Y. LeCun and Y. Bengio. Convolutional networks for images, speech, and time-series. 
-%    %In M. A. Arbib, 		editor, The Handbook of Brain Theory and Neural Networks. MIT Press, 1995
-%    
-%    % Paul Ekmans 1969 research paper "Facial Expressions"
-%
-% 
-%\end{thebibliography}
 

### 2018-04-30 doc/paper.tex
move figure

@@
 
+The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so faulty labels are likely a problem.
+It is expected that every subject shows only a single emotion, where in reality faces can express at least two emtions simultaniously (e.g happiness and surprise).
+To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft worked on re-labeling the dataset \cite{barsoum16}: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
+Instead of single class labels, probability distributions of emotion classes describe each persons facial expression.
+
 \begin{figure}[H]
@@
 \end{figure}
-
-The class labels were crowd-sourced with Amazon Mechanical Turk, where quality work is not incentivized, so faulty labels are likely a problem.
-It is expected that every subject shows only a single emotion, where in reality faces can express at least two emtions simultaniously (e.g happiness and surprise).
-To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft worked on re-labeling the dataset \cite{barsoum16}: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
-Instead of single class labels, probability distributions of emotion classes describe each persons facial expression.
 
@@
 \subsection{Training Procedure}
-The CNN is trained on a subset of the FER-dataset according to the original train/test/validation split with the FERplus prabability distributions as labels.
+The CNN is trained on a subset of the FER-dataset according to the original train-test-validation split with the FERplus prabability distributions as labels.
 The models filters and weights/biases of the fully connected layers are fit to the training set using a backpropagation algorithm.
@@
 In order to balance the training dataset, data augmentation could be used \cite{krizhevsky12}: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
-Given more time for this project, I would like to have experimented with such techniques of data augmentation.
+Given more time for this project, I would like to have experimented with such techniques of data augmentation and also with  more complex model architectures.
 

### 2018-04-30 doc/paper.tex
phrasing

@@
 Some researchers have attempted to derive emotions from detecting individual muscle movements in images of faces by identifying muscles and their degree of activation \cite{lien98}.
-However, manually extracting these relevant features from the training data requires expert knowledge or complex algorithms.
-In Convolutional Neural Networks (CNNs), features can be learned from the training data implicitly and only the image pixels are required as input for predictions.
-With this convenience, CNNs have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
+However, manually extracting the muscle activations from a set of images requires expert knowledge or complex algorithms.
+With Convolutional Neural Networks (CNNs), training data only consists of images along with matching class labels that describe the images.
+Relevant features, such as a smile that implies happiness, are learned implicitly and do not require manual extraction.
+CNNs have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
 %------------------------------------------------
@@
 The final model scored an accuracy of ~75\% on a test-set of 3557 samples.
-In live predictions with a webcam, the classifier has shown to be accurate.
+In live predictions with a webcam, the classifier has shown to be accurate in most of the predictions.
 However, predictions for the classes $disgust$ and $fear$ are hard to reproduce.
 This is most likely because the seven classes of emotion in the dataset are highly imbalanced.
-For instance disgust only has 500 samples while happy has much more.
+For instance disgust only has about 500 samples while happy has ~7000.
 This results in biased predictions with higher probabilities of the majority classes.

### 2018-05-08 doc/paper.tex
revising the paper

@@
 CNNs have been used successfully for various image recognition tasks, including facial expression recognition \cite{dumas01}.
+The aim of this work was to test how well multiple CNN models, trained on different datasets, perform together on the task of recognizing facial expressions.
 %------------------------------------------------
@@
 \section{Convolutional Neural Networks (CNNs)}
-Convolutional Neural Networks are are a special kind of multi-layer artificial neural networks.
+Convolutional Neural Networks are a special kind of multi-layer artificial neural networks.
 CNNs have been around since the 1990s \cite{lecun98}, however they have lead to breakthrough results in practical application with computational power becoming more affordable in the last years.
@@
 \subsection{Training Procedure}
-The CNN is trained on a subset of the FER-dataset according to the original train-test-validation split with the FERplus prabability distributions as labels.
-The models filters and weights/biases of the fully connected layers are fit to the training set using a backpropagation algorithm.
+The CNN is trained on a subset of 28221 training examples from the FER-dataset according to the original train-test-validation split.
+The FERplus prabability distributions are used as as labels.
+In the training subset, only images with a maximum label class probability of $> 0.5$ were considered.
+The CNN models filters and weights/biases of the fully connected layers are fit to the training set using the backpropagation algorithm.
 Therefore, a cross-entropy loss function is used to measure the model error and the Adam optimization algorithm with an initial step size of 0.001 is used to minimize the loss.
-To prevent overfitting the model on the training set, the loss value on the validation set (unseen data) is computed periodically during training.
-When the validation loss stops decreasing, training is stopped and the final model accuracy is computed on the test subset.
+To prevent overfitting the model on the training set, the loss value on the validation set (3532 unseen images) is computed periodically during training.
+When the validation loss stops decreasing, training is stopped and the final model accuracy is computed on a test subset.
+
+%------------------------------------------------
+
+\section{Results}
+The final CNN model scored an accuracy of 73.2\% on a test-set of 3482 face examples.
+Training was executed on a shared Tesla K80 GPU instance by Google Colab \cite{colab}, a cloud powered, interactive Jupyter notebook environment.
+Training with 24 iterations over the training set (epochs) took about 15 minutes.
+
+\begin{figure}[H]
+	\includegraphics[width=0.48\textwidth]{confusion}
+	\caption{Confusion matrix showing the accuracy for each class}
+\end{figure}
+
+Occlusion bla bla
+
+\begin{figure}[H]
+	\includegraphics[width=0.48\textwidth]{errors}
+	\caption{Examples of wrongly classified images}
+\end{figure}
 
@@
 % - CNN robustness to occlusion, distortions and geometric transformations.
-The final model scored an accuracy of ~75\% on a test-set of 3557 samples.
 In live predictions with a webcam, the classifier has shown to be accurate in most of the predictions.
@@
 This is most likely because the seven classes of emotion in the dataset are highly imbalanced.
-For instance disgust only has about 500 samples while happy has ~7000.
+For instance $disgust$ only has about 500 training samples while $happy$ has about 7000.
 This results in biased predictions with higher probabilities of the majority classes.
-In order to balance the training dataset, data augmentation could be used \cite{krizhevsky12}: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
+In order to balance the training dataset, data augmentation \cite{krizhevsky12} could be used: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
 Given more time for this project, I would like to have experimented with such techniques of data augmentation and also with more complex model architectures.

### 2018-05-08 doc/paper.tex
revision

@@
 \usepackage{cite} % BibTex
+\usepackage[inline]{enumitem} % enumerate
 
@@
 Automatically recognizing facial expressions is an interesting and challenging problem in the field of human computer interaction.
-Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures: anger, disgust, fear, happiness, sadness, and surprise \cite{ekman93}.
-This projects goal is to develop a classifier for real-time detection of these facial expressions using deep learning.
+Facial expressions can be interpreted and categorized into six classes of basic emotion that seem to be found across all human cultures \cite{ekman93}:
+\begin{enumerate*}[]
+\item anger
+\item disgust
+\item fear
+\item happiness
+\item sadness and
+\item surprise
+\end{enumerate*}. This projects goal is to develop a classifier for real-time detection of these facial expressions using deep learning.
 
@@
 It is expected that every subject shows only a single emotion, where in reality faces can express at least two emtions simultaniously (e.g happiness and surprise).
-To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft worked on re-labeling the dataset \cite{barsoum16}: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
+To resolve this problem and improve the overall quality of the FER-2013 dataset, a research group at Microsoft created the FERPlus extension \cite{barsoum16} with new labels for the FER-2013 dataset: To get a better approximation of the ground-truth emotions, ten workers on Mechanical Turk were asked to label each image.
 Instead of single class labels, probability distributions of emotion classes describe each persons facial expression.
@@
 	\includegraphics[width=0.48\textwidth]{ferplus_distribution}
-	\caption{FER-2013 dataset class distribution}
+	\caption{FERPlus dataset class distribution}
 \end{figure}
@@
 The FERplus prabability distributions are used as as labels.
-In the training subset, only images with a maximum label class probability of $> 0.5$ were considered.
+In the training subset, only images with a label class probability of $> 0.5$ were considered.
 The CNN models filters and weights/biases of the fully connected layers are fit to the training set using the backpropagation algorithm.
@@
 	\caption{Confusion matrix showing the accuracy for each class}
+	\label{fig::confusion}
 \end{figure}
 
-Occlusion bla bla
+The classes 3=$happy$ and 6=$neutral$ have a high rate of correctly being identified according to the confusion matrix, whereas 1=$disgust$ and 2=$fear$ are barely recognized.
 
@@
 In order to balance the training dataset, data augmentation \cite{krizhevsky12} could be used: To increase the number of training examples for a given class, the images could be flipped horizontally or rotated in multiple random angles.
-Given more time for this project, I would like to have experimented with such techniques of data augmentation and also with more complex model architectures.
+
+The examples of wrongly classified images show that especially when facial expressions are not exaggerated or posed, they can be really hard to distinguish and recognize.
+When trained with a relatively small dataset, CNNs don't seem to be robust against occulsion and geometric transformations.
+
+Given more time for this project, I would like to have experimented with techniques of data augmentation and also with more complex model architectures.
 
